---
title: "Multi-Omic Biomarker Discovery with costablr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Omic Biomarker Discovery with costablr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
set.seed(42)
```

## Overview

A multi-omic dataset is a set of partial views.  One assay sees immune-cell
state, another sees circulating proteins, and the phenotype is the same person
seen through both.  The statistical question is not only "which features are
stable?" but also "does a feature remain useful when another view is allowed to
speak?"

This vignette uses a real Onset of Labor (OOL) dataset bundled with `costablr`.
It contains two omic layers, CyTOF and Proteomics, measured in pregnant women.
The outcome is days-to-onset-of-labor (DOS), a continuous regression target.
The settings are intentionally bounded for package builds while preserving the
shape of a real train/validation analysis.

We will:

1. Loading the bundled example data with `load_ool_data()`.
2. Running per-omic STABL --- a useful baseline before integration.
3. Running the integrated multi-omic workflow with **early fusion** and
   **late fusion**.
4. Comparing selected features across strategies.
5. Exporting results.

## Learning objectives

By the end, you should be able to:

1. Understand the data structure returned by `load_ool_data()`.
2. Run STABL independently on each omic and interpret the outputs.
3. Explain the difference between early fusion and late fusion.
4. Run `stabl_multiomic_train_validate()` and navigate its result list.
5. Compare which features appear across integration strategies.
6. Export results for downstream use.

```{r load}
library(costablr)
```

---

## 1. Load the data

`load_ool_data()` reads the bundled OOL subset directly from the package.  No
file paths are needed.  It returns a named list with two elements:

- **`$x_list`**: a named list of feature matrices, one per omic layer.  Each
  matrix has samples in rows and features in columns, with matching row names.
- **`$y`**: a named numeric vector of outcomes (DOS in days), aligned to the
  rows of every matrix in `$x_list`.

**Relation to the Python tutorial notebook:** this is a bounded package
example, not a full reproduction of `Notebook examples/Tutorial Notebook.ipynb`.
The notebook uses the full OOL-CV training files (`CyTOF`, `Proteomics`, and
`Metabolomics`) and the COVID-19 proteomics classification dataset.  It also
applies a preprocessing pipeline before STABL: zero-variance filtering,
high-missingness feature filtering, median imputation, and standardisation.
Here, `load_ool_data()` ships a compact 100-feature OOL subset for `CyTOF` and
`Proteomics` only, already aligned and without missing values.  Use
`vignette("costablr-python-parity")` when the goal is to mirror the tutorial
notebook's datasets and preprocessing choices.

```{r load-data}
ool_train <- load_ool_data(split = "train")
ool_valid <- load_ool_data(split = "valid")

# Training set dimensions
lapply(ool_train$x_list, dim)
length(ool_train$y)

# Validation set dimensions
lapply(ool_valid$x_list, dim)
```

The `lapply(..., dim)` output confirms the basic geometry: the omics share rows
but differ in columns.  Row names must match across views.  `costablr` checks this
alignment and stops early if it detects a mismatch.

```{r data-summary}
# Outcome distribution (DOS = days before onset of labor)
summary(ool_train$y)
```

The `summary()` output shows the spread of DOS.  A wide range is expected; a
strong skew or narrow range changes how much prediction signal one should
expect from molecular measurements.

---

## 2. Per-omic STABL

Before integrating views, it is worth listening to each one alone.  A per-omic
fit is both a sanity check and a baseline:

1. **Sanity check** --- does each omic contain any signal at all?  If STABL
   selects zero features from an omic, that layer may not contribute useful
   information to the integrated analysis.
2. **Baseline** --- the per-omic biomarker lists can be compared to the
   integrated results to see which features survive the joint analysis.

We use compact 15-point lambda grids and bounded bootstrap counts throughout
this vignette.  Increase to 500-1000 bootstraps and broader grids for
publication-quality results.

When each fit prints, look first at the number of selected features and the FDP
estimate.  The ideal result is not simply a long list.  It is a compact list
whose stability is difficult for decoys to explain.

```{r lambda-cytof}
lambda_cytof <- auto_lambda_grid(
  ool_train$x_list$cytof, ool_train$y,
  family = "gaussian", n_lambda = 15
)
```

```{r fit-cytof}
fit_cytof <- stabl_fit(
  x               = ool_train$x_list$cytof,
  y               = ool_train$y,
  lambda_grid     = lambda_cytof,
  family          = "gaussian",
  n_bootstraps    = 200L,
  artificial_type = "modelx_knockoff",
  random_state    = 42L
)
fit_cytof
```

```{r fit-proteomics}
lambda_prot <- auto_lambda_grid(
  ool_train$x_list$proteomics, ool_train$y,
  family = "gaussian", n_lambda = 15
)

fit_prot <- stabl_fit(
  x               = scale(ool_train$x_list$proteomics),
  y               = ool_train$y,
  lambda_grid     = lambda_prot,
  family          = "gaussian",
  n_bootstraps    = 500L,
  artificial_type = "modelx_knockoff",
  random_state    = 42L
)
fit_prot
```

`get_support()` returns the features that passed the stability threshold in
each omic.  Non-overlapping lists suggest view-specific information.  Overlap,
when the same molecule is measured in both layers, is stronger evidence because
the signal has appeared twice through different measurement systems.

```{r per-omic-support}
cat("CyTOF selected features:
")
print(get_support(fit_cytof))

cat("
Proteomics selected features:
")
print(get_support(fit_prot))
```

In the stability path, each line is one feature.  Features that hold high
selection frequencies across a moderate penalty range are the most convincing.
Features that appear only when the penalty is very weak are easier to dismiss
as opportunistic.

```{r plot-per-omic, fig.alt = "Stability paths for CyTOF and Proteomics"}
plot_stabl_path(fit_cytof,     title = "CyTOF - stability path")
plot_stabl_path(fit_prot,      title = "Proteomics - stability path")
```

---

## 3. Integrated multi-omic workflow

`stabl_multiomic_train_validate()` puts the train/validation analysis into one
call.  It accepts matched omic matrices and evaluates two complementary ways to
let the views interact:

| Strategy | How it works | Best when |
|---|---|---|
| **Early fusion** | Concatenates all omic matrices into one wide matrix, then runs a *single* STABL fit across all features jointly. | Omics are measured on the same samples; you want a unified biomarker list. |
| **Late fusion** | Fits one STABL + downstream model per omic separately, then learns the optimal weighted combination of predictions on the validation set. | You want to know each omic's independent predictive contribution. |

Key arguments:

| Argument | Purpose |
|---|---|
| `x_train_list` / `x_valid_list` | Named lists of feature matrices; names must match between train and valid. |
| `lambda_grid` | Named list of lambda grids, one per omic (names must match `x_train_list`). |
| `early_fusion` / `late_fusion` | Toggle each strategy on or off independently. |
| `n_iter_lf` | Number of optimisation iterations for the late-fusion weight search. |

The printed `multi_fit` object summarises every active strategy and reports
validation metrics, Spearman rho for regression and AUROC for classification,
so the comparison is visible before digging into list elements.

```{r lambda-grid}
# Build per-omic lambda grids
lambda_list <- list(
  cytof      = auto_lambda_grid(ool_train$x_list$cytof,
                                ool_train$y, family = "gaussian", n_lambda = 15),
  proteomics = auto_lambda_grid(ool_train$x_list$proteomics,
                                ool_train$y, family = "gaussian", n_lambda = 15)
)
```

```{r multiomic-fit}
multi_fit <- stabl_multiomic_train_validate(
  x_train_list    = ool_train$x_list,
  y_train         = ool_train$y,
  lambda_grid     = lambda_list,
  x_valid_list    = ool_valid$x_list,
  y_valid         = ool_valid$y,
  family          = "gaussian",
  n_bootstraps    = 200L,
  artificial_type = "modelx_knockoff",
  early_fusion    = TRUE,
  late_fusion     = TRUE,
  n_iter_lf       = 500L,
  random_state    = 42L
)

multi_fit
```

---

## 4. Exploring results

`multi_fit` is a list because the workflow has several branches.  The main
elements are:

- **`$fits`**: named list of `stabl_fit` objects, one per omic.  These are the
  per-omic STABL results produced *within* the joint pipeline.
- **`$early_fusion`**: result of the early-fusion step (if
  `early_fusion = TRUE`), containing `$fit` (a `stabl_fit` on the concatenated
  matrix).
- **`$late_fusion`**: result of the late-fusion step (if `late_fusion = TRUE`),
  containing `$valid_predictions` (validation predictions) and `$weights`
  (per-omic importance weights).

```{r per-omic-results}
# Per-omic selected features from the joint pipeline
cat("CyTOF (from multi-omic pipeline):
")
print(get_support(multi_fit$fits$cytof))

cat("
Proteomics (from multi-omic pipeline):
")
print(get_support(multi_fit$fits$proteomics))
```

```{r early-fusion-results}
# Early fusion: joint selection across all omics simultaneously
if (!is.null(multi_fit$early_fusion)) {
  cat("Early fusion selected features:\n")
  print(get_support(multi_fit$early_fusion$fit))
}
```

Early fusion can disagree with the per-omic lists because the joint model sees
all features at once.  A feature that is useful alone may become redundant when
another omic carries the same information.  That is a property of the joint
question, not a failure of the method.

```{r late-fusion-results}
# Late fusion: optimal linear combination of per-omic predictions
if (!is.null(multi_fit$late_fusion)) {
  cat("Late fusion validation predictions (first 6):\n")
  print(head(multi_fit$late_fusion$valid_predictions))
}
```

`$valid_predictions` is a numeric vector of predicted DOS values for the held-out
validation set, one prediction per sample.  Compare these to `ool_valid$y` to
assess predictive performance.  The `$weights` element shows how much each omic
contributed to the combined prediction; a weight near zero means that view
added little beyond the other view.

```{r stability-paths, fig.alt = "CyTOF stability path from multi-omic fit"}
plot_stabl_path(multi_fit$fits$cytof,
                title = "CyTOF (multi-omic pipeline)")
plot_stabl_path(multi_fit$fits$proteomics,
                title = "Proteomics (multi-omic pipeline)")
```

---

## 5. Export results

`save_stabl_results()` writes the result of a single `stabl_fit` object to
disk:

| File | Content |
|---|---|
| `stability_scores.csv` | Feature names and their stability scores. |
| `selected_features.csv` | Only the features above the threshold. |
| `stability_path.pdf` | The stability path plot. |
| `fdr_graph.pdf` | The FDR graph. |

Set `override = TRUE` to overwrite an existing directory, which is useful when
an analysis is being rerun with more bootstraps or a revised grid.

```{r export}
out_dir <- file.path(tempdir(), "costablr_ool_cytof")

save_stabl_results(
  object    = multi_fit$fits$cytof,
  path      = out_dir,
  x         = ool_train$x_list$cytof,
  y         = ool_train$y,
  task_type = "regression",
  override  = TRUE
)

list.files(out_dir, recursive = TRUE)
```

---

## Next steps

- Use `stabl_multiomic_cv()` for a fully cross-validated evaluation when no
  pre-defined validation split is available.
- Pass `groups` to respect subject-level grouping in bootstrap resampling and
  CV fold construction.
- See `vignette("costablr-intro")` for single-omic usage and alternative learners.
