---
title: "TCGA Breast Cancer Nested-CV Head-to-Head: costablr vs DIABLO"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCGA Breast Cancer Nested-CV Head-to-Head: costablr vs DIABLO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  cache = FALSE
)
```

## Overview

The compact TCGA vignette shows the workflow.  This vignette asks for evidence.
It compares `costablr` and mixOmics DIABLO on the same three-class PAM50 breast
cancer task: Basal, Her2, and LumA.

The design is nested cross-validation.  Outer folds estimate generalisation.
Inner folds tune the models.  Feature selection is repeated inside each
training split, rather than performed once on the full dataset before
validation.  That distinction is the difference between a benchmark and an
optimistic story.

The head-to-head uses mRNA and miRNA only.  The protein block in
`mixOmics::breast.TCGA` is available only for the original training samples,
while all 220 samples have mRNA and miRNA.  Excluding protein keeps both
methods on the same samples and blocks.

The heavy computation is not run while rendering this vignette.  It is executed
through SLURM and cached as an RDS file.

## Methodological Audit

This workflow was checked against the `nestedcv` vignette guidance for p >> n
biomedical prediction.  No filter or feature-selection result is computed on
the full dataset before cross-validation.  STABL selection, downstream
multinomial refitting, DIABLO component tuning, and DIABLO `keepX` tuning all
occur inside the relevant training split before a held-out outer fold is
predicted.

The design also follows p >> n variable-selection practice:

- predictors are passed as numeric matrices and model scaling is left inside the
  fitting procedures rather than learned globally before CV;
- the outer loop estimates performance, while the inner loop selects candidates
  and tuning parameters;
- costablr uses model-X knockoff artificial features for FDP-aware stability selection;
- the deliverable is a recurrence profile over outer folds, not a single
  definitive biomarker list;
- no naive post-selection p-values are reported.

The benchmark is still exploratory.  Confirmatory extensions would include
external cohort validation and refitting unpenalised subtype models on only the
most recurrent features, with explicit post-selection qualification.

## Run on SLURM

```{r cache-setup}
analysis_script <- file.path("..", "inst", "analysis", "run_tcga_nestedcv.R")
if (!file.exists(analysis_script)) {
  analysis_script <- file.path("inst", "analysis", "run_tcga_nestedcv.R")
}
source(analysis_script)

cache_path <- file.path("..", "inst", "analysis", "cache", "tcga_nestedcv_results.rds")
if (!file.exists(cache_path)) {
  cache_path <- file.path("inst", "analysis", "cache", "tcga_nestedcv_results.rds")
}

slurm_cmd <- "sbatch inst/analysis/tcga_nestedcv.slurm"
cat("Expected cache:", cache_path, "\n")
cat("Submit command:", slurm_cmd, "\n")
```

```{r cache-status}
has_results <- tcga_cache_complete(cache_path)
if (!has_results) {
  cat(
    "Nested-CV cache is not available yet.\n",
    "Submit the SLURM job above, then render this notebook again.\n",
    sep = ""
  )
}
```

## Analysis Design

The cache records the exact settings used by the SLURM runner:

```{r design-summary, eval = has_results}
res <- readRDS(cache_path)
str(res$settings)
```

`costablr` uses multinomial stability selection with model-X knockoff artificial
features.  The candidate workflows are mRNA-only, miRNA-only, and early fusion
across mRNA and miRNA.  Candidate choice is based on inner-CV balanced error
rate.

DIABLO uses the mixOmics multiblock sPLS-DA workflow: `block.plsda()` and
`perf()` choose the number of components, `tune.block.splsda()` tunes `keepX`,
and the locked outer-fold model is evaluated with weighted-vote
`centroids.dist` predictions.

## Predictive Performance

The performance table is the first readout.  The confusion matrices show where
the methods make their mistakes, which is often more informative than a single
accuracy number.

```{r performance, eval = has_results}
res$performance
```

```{r confusion, eval = has_results}
stabl_conf <- table(
  Truth = res$costablr$outer_predictions$truth,
  Predicted = res$costablr$outer_predictions$predicted
)
diablo_conf <- table(
  Truth = res$diablo$outer_predictions$truth,
  Predicted = res$diablo$outer_predictions$predicted
)

cat("costablr confusion matrix:\n")
print(stabl_conf)
cat("\nDIABLO confusion matrix:\n")
print(diablo_conf)
```

```{r paired-fold-performance, eval = has_results}
fold_metric <- function(pred) {
  by_fold <- split(pred, pred$fold_id)
  do.call(rbind, lapply(names(by_fold), function(id) {
    m <- tcga_metrics(by_fold[[id]]$truth, by_fold[[id]]$predicted)
    data.frame(fold_id = id, m, stringsAsFactors = FALSE)
  }))
}

stabl_fold <- fold_metric(res$costablr$outer_predictions)
diablo_fold <- fold_metric(res$diablo$outer_predictions)
paired <- merge(stabl_fold, diablo_fold, by = "fold_id",
                suffixes = c("_costablr", "_DIABLO"))
paired$delta_BER <- paired$balanced_error_rate_costablr -
  paired$balanced_error_rate_DIABLO
paired$delta_accuracy <- paired$accuracy_costablr - paired$accuracy_DIABLO
summary(paired[, c("delta_BER", "delta_accuracy")])
```

## Feature Comparison

Feature recurrence is the biomarker-side counterpart to predictive
performance.  A recurrent feature is not merely selected once; it survives
across many outer training splits.

```{r feature-overlap, eval = has_results}
res$feature_comparison$overlap
```

```{r feature-recurrence, eval = has_results}
recurrence <- res$feature_comparison$recurrence
head(recurrence[order(recurrence$block, -recurrence$recurrence), ], 30)
```

```{r shared-features, eval = has_results}
shared <- do.call(rbind, lapply(split(recurrence, recurrence$block), function(tab) {
  stabl_feats <- tab$feature[tab$method == "costablr"]
  diablo_feats <- tab$feature[tab$method == "DIABLO"]
  tab[tab$feature %in% intersect(stabl_feats, diablo_feats), ]
}))
shared[order(shared$block, shared$feature, shared$method), ]
```

## Exported Artifacts

The SLURM runner writes these files next to the cache:

```{r exported-files, eval = has_results}
list.files(dirname(cache_path), pattern = "tcga_nestedcv", full.names = TRUE)
```

The CSV files are intended for manuscript tables and supplement material:
performance, feature recurrence, and feature overlap by omic block.

## Session Info

```{r session-info, eval = has_results}
res$session_info
```
