---
title: "Multi-Omic Biomarker Discovery with costablr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Omic Biomarker Discovery with costablr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
set.seed(42)
```

## Overview

A multi-omic dataset is a set of partial views.  One assay sees immune-cell
state, another sees circulating proteins, and the phenotype is the same person
seen through both.  The statistical question is not only "which features are
stable?" but also "does a feature remain useful when another view is allowed to
speak?"

This vignette uses a real Onset of Labor (OOL) dataset bundled with `costablr`.
It contains two omic layers, CyTOF and Proteomics, measured in pregnant women.
The outcome is days-to-onset-of-labor (DOS), a continuous regression target.
The settings are intentionally bounded for package builds while preserving the
shape of a real train/validation analysis.

We will:

1. Load the bundled example data with `load_ool_data()`.
2. Run per-omic STABL, a useful baseline before integration.
3. Run the integrated multi-omic workflow with **early fusion**,
   canonical **late fusion**, **STABL-selected late fusion**, and
   **Multi-Omic STABL**.
4. Compare selected features across strategies.
5. Export results.

## Learning objectives

By the end, you should be able to:

1. Understand the data structure returned by `load_ool_data()`.
2. Run STABL independently on each omic and interpret the outputs.
3. Explain the difference between early fusion, canonical late fusion, and
   STABL-selected late fusion.
4. Run explicit per-omic STABL selection, then choose downstream fusion
   methods, including Multi-Omic STABL, from the resulting object.
5. Compare which features appear across integration strategies.
6. Export results for downstream use.

```{r load}
library(costablr)
```

---

## 1. Load the data

`load_ool_data()` reads the bundled OOL subset directly from the package.  No
file paths are needed.  It returns a named list with two elements:

- **`$x_list`**: a named list of feature matrices, one per omic layer.  Each
  matrix has samples in rows and features in columns, with matching row names.
- **`$y`**: a named numeric vector of outcomes (DOS in days), aligned to the
  rows of every matrix in `$x_list`.

**Relation to the Python tutorial notebook:** this is a bounded package
example, not a full reproduction of `Notebook examples/Tutorial Notebook.ipynb`.
The notebook uses the full OOL-CV training files (`CyTOF`, `Proteomics`, and
`Metabolomics`) and the COVID-19 proteomics classification dataset.  It also
applies a preprocessing pipeline before STABL: zero-variance filtering,
high-missingness feature filtering, median imputation, and standardisation.
Here, `load_ool_data()` ships a compact 100-feature OOL subset for `CyTOF` and
`Proteomics` only, already aligned and without missing values.  Use
`vignette("costablr-python-parity")` when the goal is to mirror the tutorial
notebook's datasets and preprocessing choices.

```{r load-data}
ool_train <- load_ool_data(split = "train")
ool_valid <- load_ool_data(split = "valid")

# Training set dimensions
lapply(ool_train$x_list, dim)
length(ool_train$y)

# Validation set dimensions
lapply(ool_valid$x_list, dim)
```

The `lapply(..., dim)` output confirms the basic geometry: the omics share rows
but differ in columns.  Row names must match across views.  `costablr` checks this
alignment and stops early if it detects a mismatch.

```{r data-summary}
# Outcome distribution (DOS = days before onset of labor)
summary(ool_train$y)
```

The `summary()` output shows the spread of DOS.  A wide range is expected; a
strong skew or narrow range changes how much prediction signal one should
expect from molecular measurements.

Before building lambda grids, standardize each omic from the training split and
apply the same transformation to validation. This keeps the automatic lambda
grids on a comparable scale without learning anything from the validation
outcomes.

```{r standardize-omics}
scale_train_valid <- function(x_train, x_valid) {
  center <- colMeans(x_train)
  scale <- apply(x_train, 2, stats::sd)
  scale[!is.finite(scale) | scale == 0] <- 1

  list(
    train = sweep(sweep(x_train, 2, center), 2, scale, "/"),
    valid = sweep(sweep(x_valid, 2, center), 2, scale, "/")
  )
}

scaled_omics <- lapply(names(ool_train$x_list), function(omic) {
  scale_train_valid(ool_train$x_list[[omic]], ool_valid$x_list[[omic]])
})
names(scaled_omics) <- names(ool_train$x_list)

x_train_list <- lapply(scaled_omics, `[[`, "train")
x_valid_list <- lapply(scaled_omics, `[[`, "valid")
```

---

## 2. Per-omic STABL

Before integrating views, it is worth listening to each one alone.  A per-omic
fit is both a sanity check and a baseline:

1. **Sanity check** --- does each omic contain any signal at all?  If STABL
   selects zero features from an omic, that layer may not contribute useful
   information to the integrated analysis.
2. **Baseline** --- the per-omic biomarker lists can be compared to the
   integrated results to see which features survive the joint analysis.

We use compact 15-point lambda grids and bounded bootstrap counts throughout
this vignette.  Increase to 500-1000 bootstraps and broader grids for
publication-quality results.

When each fit prints, look first at the number of selected features and the FDP
estimate.  The ideal result is not simply a long list.  It is a compact list
whose stability is difficult for decoys to explain.

```{r lambda-cytof}
lambda_cytof <- auto_lambda_grid(
  x_train_list$cytof, ool_train$y,
  family = "gaussian", n_lambda = 15
)
```

```{r fit-cytof}
fit_cytof <- stabl_fit(
  x               = x_train_list$cytof,
  y               = ool_train$y,
  lambda_grid     = lambda_cytof,
  family          = "gaussian",
  n_bootstraps    = 200L,
  artificial_type = "modelx_knockoff",
  random_state    = 42L
)
fit_cytof
```

```{r fit-proteomics}
lambda_prot <- auto_lambda_grid(
  x_train_list$proteomics, ool_train$y,
  family = "gaussian", n_lambda = 15
)

fit_prot <- stabl_fit(
  x               = x_train_list$proteomics,
  y               = ool_train$y,
  lambda_grid     = lambda_prot,
  family          = "gaussian",
  n_bootstraps    = 200L,
  artificial_type = "modelx_knockoff",
  random_state    = 42L
)
fit_prot
```

`get_support()` returns the features that passed the stability threshold in
each omic.  Non-overlapping lists suggest view-specific information.  Overlap,
when the same molecule is measured in both layers, is stronger evidence because
the signal has appeared twice through different measurement systems.

```{r per-omic-support}
cat("CyTOF selected features:\n")
print(get_support(fit_cytof))

cat("\nProteomics selected features:\n")
print(get_support(fit_prot))
```

In the stability path, each line is one feature.  Features that hold high
selection frequencies across a moderate penalty range are the most convincing.
Features that appear only when the penalty is very weak are easier to dismiss
as opportunistic.

```{r plot-per-omic, fig.alt = "Stability paths for CyTOF and Proteomics"}
plot_stabl_path(fit_cytof,     title = "CyTOF - stability path")
plot_stabl_path(fit_prot,      title = "Proteomics - stability path")
```

---

## 3. Integrated multi-omic workflow

For STABL-selected integration, the preferred API is explicit: first create a
`stabl_per_omic()` artifact, then pass that artifact to
`stabl_late_fusion()` or `stabl_multiomics()`. This keeps the per-omic
selection boundary visible and makes it harder to accidentally reuse a
full-data selection inside cross-validation.

`stabl_multiomic_train_validate()` remains a convenient orchestration wrapper
for train/validation comparisons and for raw baseline branches such as Early
Fusion and canonical Late Fusion. It accepts matched omic matrices and evaluates
complementary ways to let the views interact:

| Strategy | How it works | Best when |
|---|---|---|
| **Early fusion** | Concatenates all omic matrices into one wide matrix, then runs a *single* STABL fit across all features jointly. | Omics are measured on the same samples; you want a unified biomarker list. |
| **Late fusion** | Fits one penalized predictive model per omic on the full per-view matrix, then learns a weighted combination of per-omic predictions. | You want the canonical prediction-level fusion baseline without STABL-selected features. |
| **STABL-selected late fusion** | Fits one STABL + downstream model per omic separately, then learns a weighted combination of training predictions and applies those weights to validation predictions. | You want to know each omic's independent predictive contribution after per-view selection. |
| **Multi-Omic STABL** | Concatenates only the per-omic STABL-selected biomarkers, then fits one final model on that filtered layer. | You want the paper-level StablSRM multi-omic final layer. |

Key arguments:

| Argument | Purpose |
|---|---|
| `x_train_list` / `x_valid_list` | Named lists of feature matrices; names must match between train and valid. |
| `lambda_grid` | Named list of lambda grids for per-omic STABL; shared grid or `"auto"` for Early Fusion because it has one concatenated input space. |
| `stabl_per_omic()` | Preferred entry point for STABL-selected downstream methods. |
| `stabl_late_fusion(per_omic)` | STABL-selected prediction-level fusion from a per-omic selector artifact. |
| `stabl_multiomics(per_omic)` | Paper-level Multi-Omic STABL final layer from selected biomarkers. |
| `early_fusion` / `late_fusion` | Wrapper switches for raw Early Fusion and canonical prediction-level Late Fusion baselines. |
| `n_iter_stacking` | Number of optimisation iterations for prediction-level stacking weight searches. |

The printed objects report selected feature counts and validation metrics, so
the comparison is visible before digging into list elements.

```{r lambda-grid}
# Build per-omic lambda grids
lambda_list <- list(
  cytof      = auto_lambda_grid(x_train_list$cytof,
                                ool_train$y, family = "gaussian", n_lambda = 15),
  proteomics = auto_lambda_grid(x_train_list$proteomics,
                                ool_train$y, family = "gaussian", n_lambda = 15)
)

lambda_shared <- auto_lambda_grid(
  do.call(cbind, x_train_list),
  ool_train$y,
  family = "gaussian",
  n_lambda = 15
)
```

```{r per-omic-artifact}
per_omic <- stabl_per_omic(
  x_train_list    = x_train_list,
  y_train         = ool_train$y,
  lambda_grid     = lambda_list,
  x_valid_list    = x_valid_list,
  y_valid         = ool_valid$y,
  family          = "gaussian",
  n_bootstraps    = 200L,
  artificial_type = "modelx_knockoff",
  random_state    = 42L
)

per_omic
```

```{r stabl-selected-integrations}
stabl_lf <- stabl_late_fusion(
  per_omic,
  n_iter_stacking = 500L,
  random_state = 42L
)

multiomics <- stabl_multiomics(per_omic)

stabl_lf
multiomics
```

For raw baselines, use the wrapper switches. Here, Early Fusion asks what is
stable after concatenating the omic matrices before STABL selection. Canonical
Late Fusion asks how well full-view penalized predictors combine at the
prediction level, without using STABL-selected features. Because this wrapper
call includes Early Fusion, it uses the shared `lambda_shared` grid required by
the single concatenated input space.

```{r raw-baseline-fit}
baseline_fit <- stabl_multiomic_train_validate(
  x_train_list    = x_train_list,
  y_train         = ool_train$y,
  lambda_grid     = lambda_shared,
  x_valid_list    = x_valid_list,
  y_valid         = ool_valid$y,
  family          = "gaussian",
  n_bootstraps    = 80L,
  artificial_type = "modelx_knockoff",
  early_fusion    = TRUE,
  late_fusion     = TRUE,
  n_iter_stacking = 500L,
  random_state    = 42L
)

baseline_fit
```

---

## 4. Exploring results

The current workflow deliberately separates the selector object from the
downstream integration methods. The main elements are:

- **`per_omic$fits`**: named list of `stabl_fit` objects, one per omic. These
  are the independent per-omic STABL selectors.
- **`baseline_fit$early_fusion`**: result of the early-fusion step (if
  `early_fusion = TRUE`), containing `$fit` (a `stabl_fit` on the concatenated
  matrix).
- **`baseline_fit$late_fusion`**: result of canonical prediction-level Late Fusion (if
  `late_fusion = TRUE`), containing per-view penalized `$models`,
  `$valid_predictions`, and `$weights`.
- **`stabl_lf`**: STABL-Selected Late Fusion, containing `$valid_predictions`
  and per-omic `$weights`.
- **`multiomics`**: paper-level Multi-Omic STABL, containing the selected
  combined final-layer matrix, provenance `$feature_map`, final `$refit`, and
  predictions.

```{r per-omic-results}
# Per-omic selected features from the selector artifact
cat("CyTOF (per-omic STABL):\n")
print(get_support(per_omic$fits$cytof))

cat("\nProteomics (per-omic STABL):\n")
print(get_support(per_omic$fits$proteomics))
```

```{r early-fusion-results}
# Early fusion: joint selection across all omics simultaneously
if (!is.null(baseline_fit$early_fusion)) {
  cat("Early fusion selected features:\n")
  print(get_support(baseline_fit$early_fusion$fit))
}
```

Early fusion can disagree with the per-omic lists because the joint model sees
all features at once.  A feature that is useful alone may become redundant when
another omic carries the same information.  That is a property of the joint
question, not a failure of the method.

```{r late-fusion-results}
# Canonical late fusion: optimal linear combination of full-view model predictions
if (!is.null(baseline_fit$late_fusion)) {
  cat("Late-fusion omic weights:\n")
  print(baseline_fit$late_fusion$weights)
}
```

```{r stabl-selected-late-fusion-results}
# STABL-selected late fusion: optimal linear combination of per-omic predictions
cat("STABL-selected late-fusion omic weights:\n")
print(stabl_lf$weights)

cat("\nSTABL-selected late-fusion validation predictions (first 6):\n")
print(head(stabl_lf$valid_predictions))
```

```{r multiomics-results}
cat("Multi-Omic STABL final-layer features (first 10):\n")
print(head(multiomics$feature_map, 10))

cat("\nMulti-Omic STABL validation predictions (first 6):\n")
print(head(multiomics$valid_predictions))
```

`$valid_predictions` is a numeric vector of predicted DOS values for the held-out
validation set, one prediction per sample. Compare these to `ool_valid$y` to
assess predictive performance. In `stabl_lf`, the `$weights` element shows how
much each omic contributed to the combined prediction; a weight near zero means
that view added little beyond the other view. In `multiomics`, the final model
weights individual selected biomarkers from different omics in one filtered
feature layer.

```{r stability-paths, fig.alt = "CyTOF stability path from multi-omic fit"}
plot_stabl_path(per_omic$fits$cytof,
                title = "CyTOF (per-omic STABL artifact)")
plot_stabl_path(per_omic$fits$proteomics,
                title = "Proteomics (per-omic STABL artifact)")
```

---

## 5. Export results

`save_stabl_results()` writes the result of a single `stabl_fit` object to
disk:

| File | Content |
|---|---|
| `stability_scores.csv` | Feature names and their stability scores. |
| `selected_features.csv` | Only the features above the threshold. |
| `stability_path.pdf` | The stability path plot. |
| `fdr_graph.pdf` | The FDR graph. |

Set `override = TRUE` to overwrite an existing directory, which is useful when
an analysis is being rerun with more bootstraps or a revised grid.

```{r export}
out_dir <- file.path(tempdir(), "costablr_ool_cytof")

save_stabl_results(
  object    = per_omic$fits$cytof,
  path      = out_dir,
  x         = x_train_list$cytof,
  y         = ool_train$y,
  task_type = "regression",
  override  = TRUE
)

list.files(out_dir, recursive = TRUE)
```

---

## Next steps

- Use `stabl_multiomic_cv()` for a fully cross-validated evaluation when no
  pre-defined validation split is available. In a custom CV loop, create a fresh
  `stabl_per_omic()` object inside each training fold before calling
  `stabl_late_fusion()` or `stabl_multiomics()`.
- Pass `groups` to respect subject-level grouping in bootstrap resampling and
  CV fold construction.
- See `vignette("costablr-intro")` for single-omic usage and alternative learners.
