---
title: "TCGA Breast Cancer: Multi-Omic Biomarker Discovery with costablr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCGA Breast Cancer: Multi-Omic Biomarker Discovery with costablr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 6,
  fig.height = 4,
  cache      = FALSE,
  cache.path = "costablr-tcga-cache/"
)
```

## Overview

Breast cancer subtype is a molecular label, but it is also a test of
integration.  The same tumour is measured by mRNA and miRNA assays.  Some
signals should be visible in one view, some in the other, and some only when the
views are allowed to compete or cooperate.

This vignette translates the **mixOmics N-Integration Chapter 6** case study
into the `costablr` idiom.  The original mixOmics workflow uses sparse
multi-block Discriminant Analysis (sgccaDA) to classify breast cancer subtypes.
Here we ask a narrower question: which mRNA and miRNA features robustly
distinguish Basal-like tumours from all other subtypes?

The analysis uses the real `breast.TCGA` train/test split from `mixOmics`, with
compact STABL settings so the vignette stays practical.  Increase the bootstrap
count and lambda-grid size before treating the result as publication evidence.

**Dataset:** `breast.TCGA` from the `mixOmics` package.

- Training: mRNA and miRNA measurements on the same tumour samples.
- Test: held-out mRNA and miRNA measurements with matching feature sets.
- Outcome: PAM50 subtype binarised to **Basal vs. non-Basal**.
- Protein data is excluded to match the two-omic scope used here.

## Prerequisites

```{r prerequisites}
if (!requireNamespace("mixOmics", quietly = TRUE)) {
  stop(
    "This vignette requires mixOmics (Bioconductor). ",
    "Install with: BiocManager::install('mixOmics')"
  )
}
library(costablr)
library(mixOmics)
```

## 1. Load and prepare data

The train and test matrices are already aligned within the `mixOmics` object.
We extract mRNA and miRNA, then turn the PAM50 subtype into a binary Basal vs
non-Basal outcome.

```{r load-data}
data(breast.TCGA)

X_train_mrna  <- breast.TCGA$data.train$mrna
X_train_mirna <- breast.TCGA$data.train$mirna
X_test_mrna   <- breast.TCGA$data.test$mrna
X_test_mirna  <- breast.TCGA$data.test$mirna

cat("Training mRNA:  ", nrow(X_train_mrna),  "x", ncol(X_train_mrna),  "\n")
cat("Training miRNA: ", nrow(X_train_mirna), "x", ncol(X_train_mirna), "\n")
cat("Test mRNA:      ", nrow(X_test_mrna),   "x", ncol(X_test_mrna),   "\n")
cat("Test miRNA:     ", nrow(X_test_mirna),  "x", ncol(X_test_mirna),  "\n")
```

```{r outcome}
subtype_train <- breast.TCGA$data.train$subtype
subtype_test  <- breast.TCGA$data.test$subtype

y_train <- as.integer(subtype_train == "Basal")
y_test  <- as.integer(subtype_test  == "Basal")

names(y_train) <- rownames(X_train_mrna)
names(y_test)  <- rownames(X_test_mrna)

cat("Training: ", sum(y_train), "Basal /", sum(y_train == 0), "non-Basal\n")
cat("Test:     ", sum(y_test),  "Basal /", sum(y_test == 0),  "non-Basal\n")
```

## 2. Per-omic STABL

The first pass treats each assay as its own experiment.  This tells us whether
mRNA and miRNA each contain stable Basal signal before the integrated model is
allowed to mix them.  The vignette uses 30 bootstraps and 12 lambda values to
keep runtime bounded; use 500-1000 bootstraps for final reporting.

```{r mrna-fit}
lambda_mrna <- auto_lambda_grid(
  X_train_mrna, y_train,
  family   = "binomial",
  n_lambda = 12
)

fit_mrna <- stabl_fit(
  x               = X_train_mrna,
  y               = y_train,
  lambda_grid     = lambda_mrna,
  family          = "binomial",
  n_bootstraps    = 30L,
  artificial_type = "random_permutation",
  random_state    = 42L
)
fit_mrna
```

```{r mirna-fit}
lambda_mirna <- auto_lambda_grid(
  X_train_mirna, y_train,
  family   = "binomial",
  n_lambda = 12
)

fit_mirna <- stabl_fit(
  x               = X_train_mirna,
  y               = y_train,
  lambda_grid     = lambda_mirna,
  family          = "binomial",
  n_bootstraps    = 30L,
  artificial_type = "random_permutation",
  random_state    = 42L
)
fit_mirna
```

```{r selected-per-omic}
cat("mRNA selected features:\n")
print(get_support(fit_mrna))

cat("\nmiRNA selected features:\n")
print(get_support(fit_mirna))
```

```{r stability-paths, fig.alt = "mRNA and miRNA stability paths"}
print(plot_stabl_path(fit_mrna,  title = "mRNA - stability path"))
print(plot_stabl_path(fit_mirna, title = "miRNA - stability path"))
```

## 3. Integrated multi-omic pipeline

The integrated workflow repeats the per-omic fits inside a common pipeline,
adds early fusion on the concatenated matrix, and learns a STABL-selected
late-fusion validation predictor.  In other words, it asks three related questions:
what is stable in each view, what is stable when all features compete, and how
much does each view help prediction on held-out tumours?

```{r integrated-fit}
lambda_list <- list(
  mrna  = lambda_mrna,
  mirna = lambda_mirna
)

multi_fit <- stabl_multiomic_train_validate(
  x_train_list    = list(mrna = X_train_mrna,  mirna = X_train_mirna),
  x_valid_list    = list(mrna = X_test_mrna,   mirna = X_test_mirna),
  y_train         = y_train,
  y_valid         = y_test,
  lambda_grid     = lambda_list,
  family          = "binomial",
  n_bootstraps    = 30L,
  artificial_type = "random_permutation",
  random_state    = 42L,
  early_fusion    = TRUE,
  late_fusion     = TRUE,
  stabl_selected_late_fusion = TRUE,
  n_iter_stacking = 500L
)
multi_fit
```

```{r integrated-results}
cat("mRNA (integrated):\n")
print(get_support(multi_fit$fits$mrna))

cat("\nmiRNA (integrated):\n")
print(get_support(multi_fit$fits$mirna))

cat("\nEarly fusion selected features:\n")
print(get_support(multi_fit$early_fusion$fit))

cat("\nLate-fusion omic weights:\n")
print(multi_fit$late_fusion$weights)

cat("\nSTABL-selected late-fusion omic weights:\n")
print(multi_fit$stabl_selected_late_fusion$weights)
```

```{r integrated-paths, fig.alt = "Integrated mRNA and miRNA stability paths"}
print(plot_stabl_path(multi_fit$fits$mrna,  title = "Integrated - mRNA"))
print(plot_stabl_path(multi_fit$fits$mirna, title = "Integrated - miRNA"))
```

## 4. Validation performance

Feature stability is only one axis of evidence.  The held-out test set asks
whether the selected features and fusion weights carry predictive information
outside the training data.

```{r validation-performance}
lf_preds <- multi_fit$stabl_selected_late_fusion$valid_predictions

plot_roc(y_true = y_test, y_preds = lf_preds,
         title = "STABL-selected late fusion - ROC curve (test set)")
plot_prc(y_true = y_test, y_preds = lf_preds,
         title = "STABL-selected late fusion - PRC curve (test set)")

pred_class <- as.integer(lf_preds >= 0.5)
conf_mat   <- table(Predicted = pred_class, Actual = y_test)
print(conf_mat)

sensitivity <- conf_mat["1", "1"] / sum(conf_mat[, "1"])
specificity <- conf_mat["0", "0"] / sum(conf_mat[, "0"])
ber <- 1 - (sensitivity + specificity) / 2

cat(sprintf("\nSensitivity (Basal):     %.3f\n", sensitivity))
cat(sprintf("Specificity (non-Basal): %.3f\n", specificity))
cat(sprintf("Balanced Error Rate:     %.3f\n", ber))
```

```{r feature-boxplots, fig.alt = "Distribution of selected mRNA features"}
mrna_feats <- names(which(get_support(multi_fit$fits$mrna)))
if (length(mrna_feats) > 0) {
  boxplot_features(
    features = mrna_feats,
    x        = X_train_mrna,
    y        = y_train,
    title    = "Top mRNA features - training set",
    ncol     = min(3L, length(mrna_feats))
  )
}
```

## 5. Export results

For a vignette, exports go to `tempdir()` so rendering does not leave generated
files in the package source tree.  In a real analysis, these CSVs are the
bridge to notebooks, review tables, and downstream biological interpretation.

```{r export-results}
tcga_out_dir <- file.path(tempdir(), "costablr_tcga_results")
export_stabl_to_csv(multi_fit$fits$mrna,  path = file.path(tcga_out_dir, "mrna"))
export_stabl_to_csv(multi_fit$fits$mirna, path = file.path(tcga_out_dir, "mirna"))
list.files(tcga_out_dir, recursive = TRUE)
```

## 6. Comparison with mixOmics

The two toolchains ask related but not identical questions.  mixOmics builds
latent components for discrimination.  `costablr` centers the feature list and
calibrates selection stability against decoys.

| Aspect | mixOmics (sgccaDA) | costablr |
|---|---|---|
| Method | Sparse CCA + LDA | Bootstrap-stability lasso |
| Feature selection | L1 penalty tuned by CV | FDP-controlled stability threshold |
| Integration | Joint latent-space optimisation | Early, Late, STABL-selected late, or Multi-Omic STABL |
| Primary metric | BER on test set | FDP+ stability score |
| Output | Components + loadings | Feature names + stability scores |

Both approaches target compact mRNA and miRNA biomarker sets for the Basal
subtype.  STABL's advantage is explicit FDP control with no assumption that the
omics must share a latent space.

```{r session-info}
sessionInfo()
```
