---
title: "Cooperative Fusion for Multi-Omic Biomarker Discovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cooperative Fusion for Multi-Omic Biomarker Discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
set.seed(42)
```

```{r multiview-check, include = FALSE}
# Cooperative fusion requires the optional multiview package.
# If it is not installed, code chunks are shown but not evaluated.
if (!requireNamespace("multiview", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
  message(
    "The 'multiview' package is not installed. ",
    "Code chunks will be displayed but not evaluated. ",
    "Install it with: install.packages('multiview') ",
    "or R CMD INSTALL <path/to/multiview>."
  )
} else {
  library(costablr)
}
```

## Overview

Early fusion asks all features to compete in one wide matrix.  STABL-selected
late fusion lets each omic select biomarkers and make its own prediction, then
combines the predictions afterward.
Cooperative fusion asks a subtler question: can the views be fitted together
without erasing their identities?

In `costablr`, the preferred STABL-selected cooperative workflow is
`stabl_per_omic()` followed by `stabl_cooperative()`.  STABL selection remains
independent inside each Omic View, then the optional `multiview` package fits a
cooperative final layer using only the selected biomarkers. Each view keeps its
own coefficient path, but the paths are encouraged to agree through a
cooperation strength `rho`.  The omics are neither fully pooled nor fully
isolated.

**When should you use cooperative fusion?**

| Strategy | How it works | Best when |
|---|---|---|
| **Per-omic** | One STABL fit per omic, independently. | Baseline; or when views are likely orthogonal. |
| **Early fusion** | Concatenate all omics, one joint STABL fit. | Omics are tightly correlated; you want a single feature list. |
| **Late fusion** | Fit one full-view predictive model per omic, then learn a weighted blend of predictions. | You want the canonical prediction-level fusion baseline. |
| **STABL-selected late fusion** | Per-omic STABL fits and Final Refits, then learn a weighted blend of predictions. | Each omic may predict independently after selection; you want omic weights. |
| **Cooperative STABL** | Per-omic STABL selection first, then a joint multiview final layer on selected biomarkers with cooperation strength `rho`. | Omics share some signal structure; you want the final predictive layer to reward agreement without changing STABL's per-omic selection contract. |

The cooperation strength `rho` controls how much the views are encouraged to
share signal. At `rho = 0`, the agreement penalty is absent and the cooperative
fit is early-fusion-like because the view-specific predictors are optimized
through their summed prediction. Positive `rho` values penalise disagreement
between views; `rho = 1` is the standard late-fusion-like reference point in
the cooperative-learning literature. `costablr` tunes `rho` over a user-supplied
grid.

## Learning objectives

By the end, you should be able to:

1. Explain what `rho` controls in cooperative fusion and how to choose a grid.
2. Run `stabl_per_omic()` followed by `stabl_cooperative()` and navigate the
   cooperative result.
3. Read the `print()` output for cooperative fusion details.
4. Understand which `cooperation_selection` and `cooperation_selector`
   combinations are supported and why.
5. Compare cooperative fusion feature selections to early fusion and
   STABL-selected late fusion, with canonical late fusion available as a
   prediction-level baseline.
6. Keep Cooperative STABL fold-local when estimating performance by
   cross-validation.

---

## 1. Load the data

We use the same Onset of Labor (OOL) dataset as `vignette("costablr-multiomic")`:
CyTOF immune profiling and Proteomics measured on the same women, with
days-to-onset-of-labor (DOS) as the continuous outcome.  The examples use
compact grids so the vignette remains buildable; enlarge the bootstrap count
and `rho` grid for a final analysis.

```{r load}
library(costablr)
```

```{r load-data}
ool_train <- load_ool_data(split = "train")
ool_valid <- load_ool_data(split = "valid")

scale_train_valid <- function(x_train, x_valid) {
  center <- colMeans(x_train)
  scale <- apply(x_train, 2, stats::sd)
  scale[!is.finite(scale) | scale == 0] <- 1

  list(
    train = sweep(sweep(x_train, 2, center), 2, scale, "/"),
    valid = sweep(sweep(x_valid, 2, center), 2, scale, "/")
  )
}

scaled_omics <- lapply(names(ool_train$x_list), function(omic) {
  scale_train_valid(ool_train$x_list[[omic]], ool_valid$x_list[[omic]])
})
names(scaled_omics) <- names(ool_train$x_list)

x_train_list <- lapply(scaled_omics, `[[`, "train")
x_valid_list <- lapply(scaled_omics, `[[`, "valid")

# Shape of each omic layer
lapply(x_train_list, dim)
```

---

## 2. Build lambda grids

Cooperative fusion still needs a lambda grid for each omic.  We build one grid
per view with `auto_lambda_grid()` and pass the grids as a named list.  Twelve
lambda values are enough to demonstrate the workflow without making each
cooperative fit too expensive. We also create a shared grid for the raw Early
Fusion baseline, because Early Fusion has one concatenated input space.

```{r lambda-grids}
lambda_list <- list(
  cytof      = auto_lambda_grid(
    x_train_list$cytof, ool_train$y,
    family = "gaussian", n_lambda = 12
  ),
  proteomics = auto_lambda_grid(
    x_train_list$proteomics, ool_train$y,
    family = "gaussian", n_lambda = 12
  )
)

lambda_shared <- auto_lambda_grid(
  do.call(cbind, x_train_list),
  ool_train$y,
  family = "gaussian",
  n_lambda = 12
)

# Number of lambda values per omic
vapply(lambda_list, nrow, integer(1L))
```

---

## 3. Running Cooperative STABL

First build the reusable per-omic STABL artifact. This is the object boundary:
it contains the per-view selectors, selected matrices, final per-view refits,
and validation metadata. Do not reuse this object across outer CV folds; create
it inside each training fold. This compact tutorial uses a permissive
`hard_threshold` so both omic views remain available for the cooperative final
layer; final analyses should prefer the FDP+-selected threshold unless a hard
threshold is justified in the analysis plan.

```{r per-omic-fit}
per_omic <- stabl_per_omic(
  x_train_list    = x_train_list,
  y_train         = ool_train$y,
  lambda_grid     = lambda_list,
  x_valid_list    = x_valid_list,
  y_valid         = ool_valid$y,
  family          = "gaussian",
  n_bootstraps    = 30L,
  artificial_type = "random_permutation",
  hard_threshold  = 0.01,
  random_state    = 42L
)
```

Then run cooperative fusion on the STABL-selected biomarkers with a `rho` grid.
`costablr` fits one multiview model per `rho` value using shared inner CV
folds, then picks the best `rho` by the chosen `cooperation_type_measure`.

Key arguments:

| Argument | Default | Purpose |
|---|---|---|
| `rho` | `0` | Scalar or vector of cooperation strengths to try. |
| `cooperation_selection` | `"cv"` | `"cv"` (inner CV) or `"validation"` (held-out set). |
| `cooperation_selector` | `"lambda.min"` | Which lambda to pick: `"lambda.min"` or `"lambda.1se"` (CV only). |
| `cooperation_type_measure` | `"default"` | Tuning metric; defaults to `"mse"` for Gaussian. |
| `cooperation_nfolds` | `5L` | Number of inner CV folds when `cooperation_selection = "cv"`. |

```{r cooperative-fit}
cf <- stabl_cooperative(
  per_omic,
  rho                   = c(0, 0.25, 0.5),
  cooperation_selection = "cv",
  cooperation_selector  = "lambda.min",
  cooperation_nfolds    = 3L
)
```

`print()` summarizes the selected views, tuning mode, chosen `rho`, and number
of cooperative features:

```{r print-fit}
cf
```

**Reading the cooperative output:**

- **`selection`** --- how `rho` was tuned: `"cv"` here.
- **`rho (chosen)`** --- the `rho` value selected by cross-validation.  A value
  of `0` means the data did not benefit from cross-view cooperation; a larger
  value means views actively helped each other.
- **`selector`** --- which lambda rule was applied (`lambda.min`).
- **`type.measure`** --- the CV metric used for tuning (e.g., `mse`).
- **`score`** --- the CV score at the chosen `rho` and `lambda`.
- **`selected feats`** --- total number of features selected across all views.

---

## 4. Navigating the cooperative result

The `stabl_cooperative()` result contains the selected model and the tuning
surface that led to it.

```{r cf-element}
# Chosen hyperparameters
cat("Chosen rho:     ", cf$rho, "\n")
cat("Chosen lambda:  ", round(cf$selected_lambda, 4), "\n")
cat("Selector:       ", cf$selector, "\n")
cat("CV score (MSE): ", round(cf$score, 4), "\n")
```

### Selected features per view

Use `get_cooperative_features()` to read the selected cooperative features
through the public accessor:

```{r cf-selected}
cat("CyTOF features selected by cooperative fusion:\n")
print(get_cooperative_features(cf, view = "cytof"))

cat("\nProteomics features selected by cooperative fusion:\n")
print(get_cooperative_features(cf, view = "proteomics"))
```

### Predictions

`$train_predictions` and `$valid_predictions` hold the multiview predictions
from the chosen model:

```{r cf-predictions}
cat("First 6 training predictions:\n")
print(head(cf$train_predictions))

cat("\nFirst 6 validation predictions:\n")
print(head(cf$valid_predictions))
```

### Tuning diagnostics

`$diagnostics` records the CV score at every `rho` in the grid, so the tuning
choice can be inspected rather than treated as a black box:

```{r cf-diagnostics}
# One row per rho value
print(cf$diagnostics[, c("rho", "lambda", "metric_value", "selected")])
```

The `selected` column marks the winning `rho`.  If the surface is flat, with
similar `metric_value` across `rho`, the omics are not gaining much from
cooperation and `rho = 0` is a defensible choice.

---

## 5. Policy constraints

Some combinations are rejected on purpose.  These guards prevent analyses that
look reasonable in code but do not have a valid tuning interpretation.

### Constraint 1 - `lambda.1se` requires CV mode

`lambda.1se` applies the "one-standard-error" rule and is only well-defined
when cross-validation is used to estimate the error surface.  Requesting it
with `cooperation_selection = "validation"` raises an error:

```{r guard-lambda1se, error = TRUE}
stabl_cooperative(
  per_omic,
  rho                   = 0,
  cooperation_selection = "validation",
  cooperation_selector  = "lambda.1se"
)
```

### Constraint 2 - Cox family cannot use validation-based tuning

The multiview Cox path does not support external validation scoring, so
`cooperation_selection = "validation"` with `family = "cox"` is not supported.
Use `cooperation_selection = "cv"` for Cox models:

```{r guard-cox-validation, error = TRUE}
library(survival)

set.seed(42)
n <- 20L
ids <- paste0("s", seq_len(n))
x_a <- matrix(rnorm(n * 4), nrow = n,
               dimnames = list(ids, paste0("a", seq_len(4))))
x_b <- matrix(rnorm(n * 4), nrow = n,
               dimnames = list(ids, paste0("b", seq_len(4))))
y_surv <- Surv(time = rexp(n), event = rbinom(n, 1L, 0.7))
rownames(y_surv) <- ids

per_omic_cox <- stabl_per_omic(
  x_train_list          = list(omic_a = x_a, omic_b = x_b),
  y_train               = y_surv,
  x_valid_list          = list(omic_a = x_a, omic_b = x_b),
  y_valid               = y_surv,
  lambda_grid           = data.frame(lambda = 1e-5),
  family                = "cox",
  n_bootstraps          = 2L,
  artificial_type       = NULL,
  hard_threshold        = 1e-9,
  random_state          = 42L
)

stabl_cooperative(
  per_omic_cox,
  rho                   = 0,
  cooperation_selection = "validation"
)
```

---

## 6. What happens when `multiview` is not installed

Cooperative STABL is optional because its final layer uses the `multiview`
backend. If `multiview` is absent and you call `stabl_cooperative()`,
`costablr` stops immediately with an informative error before any cooperative
computation runs:

```
Error: Cooperative fusion requires the optional 'multiview' package to be installed.
```

Install the bundled version with:

```r
# From the costablr repository root:
R CMD INSTALL multiview
# or from CRAN when available:
install.packages("multiview")
```

Avoiding Cooperative STABL never triggers this check, so the rest of
`costablr` works without `multiview`.

---

## 7. Comparing Cooperative, Early, and STABL-Selected Late Fusion

The most useful comparison is often side by side.  Early Fusion is still a
raw-feature single-selector strategy, so we get it from the train/validation
wrapper. STABL-Selected Late Fusion and Cooperative STABL both consume the
same `per_omic` artifact created above:

```{r all-strategies}
early_fit <- stabl_multiomic_train_validate(
  x_train_list    = x_train_list,
  y_train         = ool_train$y,
  lambda_grid     = lambda_shared,
  x_valid_list    = x_valid_list,
  y_valid         = ool_valid$y,
  family          = "gaussian",
  n_bootstraps    = 30L,
  artificial_type = "random_permutation",
  random_state    = 42L,
  early_fusion    = TRUE
)

stabl_lf <- stabl_late_fusion(
  per_omic,
  n_iter_stacking = 250L,
  random_state = 42L
)
```

Compare feature sets across strategies:

```{r compare-features}
ef_features <- early_fit$early_fusion$selected_features
cf_features <- unique(unlist(get_cooperative_features(cf),
                             use.names = FALSE))
per_omic_features <- unique(unlist(per_omic$selected_features,
                                   use.names = FALSE))

cat("Early fusion selected:      ", length(ef_features), "features\n")
cat("Cooperative STABL selected: ", length(cf_features), "features\n")
cat("Per-omic union selected:    ", length(per_omic_features), "features\n")

cat("\nSTABL-selected late-fusion omic weights:\n")
print(stabl_lf$weights)

cat("\nIn Cooperative STABL but not early fusion:\n")
print(setdiff(cf_features, ef_features))

cat("\nIn early fusion but not Cooperative STABL:\n")
print(setdiff(ef_features, cf_features))
```

**Interpreting differences:** early fusion sees all features simultaneously in
one penalised model. Cooperative STABL first applies per-view STABL selection,
then fits per-view paths that are encouraged to agree. Features unique to
Cooperative STABL may carry cross-view signal that is diluted in the
concatenated early-fusion space.

---

## 8. Cross-validation with Cooperative STABL

For performance estimation, the key rule is that `stabl_per_omic()` must be
created inside each training fold. Do not create one `per_omic` object on all
samples and reuse it inside CV; that would leak feature-selection information
from the assessment fold into the selected biomarker set.

This multiplies cost by the number of outer folds, so the code is shown but not
evaluated during vignette builds:

```{r cv-object-shape, eval = FALSE}
fold_per_omic <- stabl_per_omic(
  x_train_list = x_train_fold,
  y_train = y_train_fold,
  lambda_grid = lambda_list,
  x_valid_list = x_valid_fold,
  y_valid = y_valid_fold,
  family = "gaussian",
  n_bootstraps = 20L,
  artificial_type = "random_permutation",
  random_state = 42L
)

fold_cf <- stabl_cooperative(
  fold_per_omic,
  rho = c(0, 0.3),
  cooperation_selection = "cv",
  cooperation_nfolds = 3L
)
```

In a full resampling loop, collect each fold's `fold_cf$valid_predictions`,
`fold_cf$rho`, and `get_cooperative_diagnostics(fold_cf)` into a fold-level
results table. Use `stabl_multiomic_cv()` for Early Fusion, canonical Late
Fusion, or other flag-driven baseline branches; use the explicit loop above
when the estimand is Cooperative STABL.

---

## 9. Choosing `rho` and design guidance

The useful `rho` grid is usually modest.  Treat it as a scientific sensitivity
analysis: did the views need to cooperate, and if so, how strongly?

| Decision | Guidance |
|---|---|
| **Grid density** | Start with `c(0, 0.1, 0.3, 0.5, 1)`. Add finer points only if the CV surface has a clear non-zero minimum. |
| **Grid upper bound** | Values above 1 are allowed. Expand the grid if the selected `rho` lands on the upper boundary or if prior biology suggests very strong agreement. |
| **`lambda.min` vs `lambda.1se`** | `lambda.1se` gives a sparser model at the cost of a higher CV error; prefer `lambda.min` when prediction is the goal. |
| **`cooperation_nfolds`** | 5 (default) is suitable for n >= 50.  Use 3 for smaller datasets.  Must be >= 3. |
| **`cooperation_selection = "validation"`** | Use when you have a true held-out set and want to tune `rho` and `lambda` together on it.  Not supported for `family = "cox"`. |
| **rho = 0 wins** | If the chosen `rho` is always 0, the views are not benefiting from cooperation.  Consider whether the omics share any biological pathway. |

---

## Next steps

- See `vignette("costablr-multiomic")` for per-omic, early, and
  STABL-selected late fusion on the same OOL dataset.
- See `vignette("costablr-intro")` for single-omic STABL and learner options.
- Pass `groups` to `stabl_per_omic()` to prevent leakage in longitudinal
  designs; use the same group split when constructing custom outer folds.
- For binomial and Cox outcomes, set `family = "binomial"` or `family = "cox"`
  and adjust `cooperation_type_measure`, such as `"auc"` for binomial or
  `"deviance"` for Cox.
