---
title: "TCGA Breast Cancer Nested-CV Head-to-Head: stablr vs DIABLO"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCGA Breast Cancer Nested-CV Head-to-Head: stablr vs DIABLO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  cache = FALSE
)
```

## Overview

This notebook is the HPC-backed research version of the compact TCGA vignette.
It compares `stablr` and mixOmics DIABLO on the same three-class PAM50 breast
cancer task: Basal, Her2, and LumA.

The analysis follows a nested cross-validation design.  The outer folds estimate
generalisation.  The inner folds perform model selection and tuning.  Feature
selection is therefore repeated inside each training split, not performed once
on the full data before validation.

The primary head-to-head uses mRNA and miRNA only.  The protein block in
`mixOmics::breast.TCGA` is available only for the original training samples,
while all 220 samples have mRNA and miRNA.  Excluding protein keeps the two
methods on exactly the same samples and blocks.

The heavy computation is not run while rendering this vignette.  It is executed
through SLURM and cached as an RDS file.

## Methodological Audit

This workflow was checked against the `nestedcv` vignette guidance for p >> n
biomedical prediction.  In particular, no filter or feature-selection result is
computed on the full dataset before cross-validation.  STABL selection,
downstream multinomial refitting, DIABLO component tuning, and DIABLO `keepX`
tuning all occur inside the relevant training split before a held-out outer
fold is predicted.

The design also follows p >> n variable-selection practice:

- predictors are passed as numeric matrices and model scaling is left inside the
  fitting procedures rather than learned globally before CV;
- the outer loop estimates performance, while the inner loop selects candidates
  and tuning parameters;
- stablr uses knockoff artificial features for FDP-aware stability selection;
- the deliverable is a recurrence profile over outer folds, not a single
  definitive biomarker list;
- no naive post-selection p-values are reported.

Potential confirmatory extensions, outside this cached benchmark, are external
cohort validation and refitting unpenalised subtype models on only the most
recurrent features to report effect estimates with appropriate post-selection
qualification.

## Run on SLURM

```{r cache-setup}
analysis_script <- file.path("..", "inst", "analysis", "run_tcga_nestedcv.R")
if (!file.exists(analysis_script)) {
  analysis_script <- file.path("r-pkg", "stablr", "inst", "analysis", "run_tcga_nestedcv.R")
}
source(analysis_script)

cache_path <- file.path("..", "inst", "analysis", "cache", "tcga_nestedcv_results.rds")
if (!file.exists(cache_path)) {
  cache_path <- file.path("r-pkg", "stablr", "inst", "analysis", "cache", "tcga_nestedcv_results.rds")
}

slurm_cmd <- "sbatch r-pkg/stablr/inst/analysis/tcga_nestedcv.slurm"
cat("Expected cache:", cache_path, "\n")
cat("Submit command:", slurm_cmd, "\n")
```

```{r cache-status}
has_results <- tcga_cache_complete(cache_path)
if (!has_results) {
  cat(
    "Nested-CV cache is not available yet.\n",
    "Submit the SLURM job above, then render this notebook again.\n",
    sep = ""
  )
}
```

## Analysis Design

```{r design-summary, eval = has_results}
res <- readRDS(cache_path)
str(res$settings)
```

`stablr` uses multinomial stability selection with knockoff artificial features.
The candidate workflows are mRNA-only, miRNA-only, and early fusion across mRNA
and miRNA.  Candidate choice is based on inner-CV balanced error rate.

DIABLO uses the mixOmics multiblock sPLS-DA workflow: `block.plsda()` and
`perf()` choose the number of components, `tune.block.splsda()` tunes `keepX`,
and the locked outer-fold model is evaluated with weighted-vote
`centroids.dist` predictions.

## Predictive Performance

```{r performance, eval = has_results}
res$performance
```

```{r confusion, eval = has_results}
stabl_conf <- table(
  Truth = res$stablr$outer_predictions$truth,
  Predicted = res$stablr$outer_predictions$predicted
)
diablo_conf <- table(
  Truth = res$diablo$outer_predictions$truth,
  Predicted = res$diablo$outer_predictions$predicted
)

cat("stablr confusion matrix:\n")
print(stabl_conf)
cat("\nDIABLO confusion matrix:\n")
print(diablo_conf)
```

```{r paired-fold-performance, eval = has_results}
fold_metric <- function(pred) {
  by_fold <- split(pred, pred$fold_id)
  do.call(rbind, lapply(names(by_fold), function(id) {
    m <- tcga_metrics(by_fold[[id]]$truth, by_fold[[id]]$predicted)
    data.frame(fold_id = id, m, stringsAsFactors = FALSE)
  }))
}

stabl_fold <- fold_metric(res$stablr$outer_predictions)
diablo_fold <- fold_metric(res$diablo$outer_predictions)
paired <- merge(stabl_fold, diablo_fold, by = "fold_id",
                suffixes = c("_stablr", "_DIABLO"))
paired$delta_BER <- paired$balanced_error_rate_stablr -
  paired$balanced_error_rate_DIABLO
paired$delta_accuracy <- paired$accuracy_stablr - paired$accuracy_DIABLO
summary(paired[, c("delta_BER", "delta_accuracy")])
```

## Feature Comparison

```{r feature-overlap, eval = has_results}
res$feature_comparison$overlap
```

```{r feature-recurrence, eval = has_results}
recurrence <- res$feature_comparison$recurrence
head(recurrence[order(recurrence$block, -recurrence$recurrence), ], 30)
```

```{r shared-features, eval = has_results}
shared <- do.call(rbind, lapply(split(recurrence, recurrence$block), function(tab) {
  stabl_feats <- tab$feature[tab$method == "stablr"]
  diablo_feats <- tab$feature[tab$method == "DIABLO"]
  tab[tab$feature %in% intersect(stabl_feats, diablo_feats), ]
}))
shared[order(shared$block, shared$feature, shared$method), ]
```

## Exported Artifacts

The SLURM runner writes these files next to the cache:

```{r exported-files, eval = has_results}
list.files(dirname(cache_path), pattern = "tcga_nestedcv", full.names = TRUE)
```

The CSV files are intended for manuscript tables and supplement material:
performance, feature recurrence, and feature overlap by omic block.

## Session Info

```{r session-info, eval = has_results}
res$session_info
```
