---
title: "Multi-Omic Biomarker Discovery with stablr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Omic Biomarker Discovery with stablr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
set.seed(42)
```

## Overview

This vignette demonstrates multi-omic biomarker discovery using the `stablr`
package on a real Onset of Labor (OOL) dataset.  The dataset contains two omic
layers (CyTOF and Proteomics) measured in 150 pregnant women.  The prediction
target is days-to-onset-of-labor (DOS), a continuous regression outcome.

**Why multi-omic?**  Single-omic analyses can miss biomarkers that are
individually weak but collectively informative.  By combining immune-cell
profiling (CyTOF) with protein abundance (Proteomics), we can ask whether each
layer contributes independent signal and how to integrate them optimally.

We cover:

1. Loading the bundled example data with `load_ool_data()`.
2. Running per-omic STABL --- a useful baseline before integration.
3. Running the integrated multi-omic workflow with **early fusion** and
   **late fusion**.
4. Comparing selected features across strategies.
5. Exporting results.

## Learning objectives

By the end of this vignette you will be able to:

1. Understand the data structure returned by `load_ool_data()`.
2. Run STABL independently on each omic and interpret the outputs.
3. Explain the difference between early fusion and late fusion.
4. Run `stabl_multiomic_train_validate()` and navigate its result list.
5. Compare which features appear across integration strategies.
6. Export results for downstream use.

```{r load}
library(stablr)
```

---

## 1. Load the data

`load_ool_data()` reads the pre-bundled OOL subset directly from the package.
No file paths are needed.  It returns a named list with two elements:

- **`$x_list`**: a named list of feature matrices, one per omic layer.  Each
  matrix has samples in rows and features in columns, with matching row names.
- **`$y`**: a named numeric vector of outcomes (DOS in days), aligned to the
  rows of every matrix in `$x_list`.

```{r load-data}
ool_train <- load_ool_data(split = "train")
ool_valid <- load_ool_data(split = "valid")

# Training set dimensions
lapply(ool_train$x_list, dim)
length(ool_train$y)

# Validation set dimensions
lapply(ool_valid$x_list, dim)
```

The `lapply(..., dim)` output confirms that CyTOF and Proteomics share the same
number of rows (samples) but differ in columns (features).  The names of rows
must match across omics --- `stablr` checks this alignment automatically and
stops with an informative error if it detects a mismatch.

```{r data-summary}
# Outcome distribution (DOS = days before onset of labor)
summary(ool_train$y)
```

The `summary()` output shows the range of DOS values.  A wide range is expected
(e.g. 0-90 days); note whether the distribution is roughly symmetric or skewed,
as this informs how predictable the outcome may be from molecular data.

---

## 2. Per-omic STABL

Before running the integrated pipeline, it is instructive to see which features
STABL selects in each omic independently.  This per-omic step serves two
purposes:

1. **Sanity check** --- does each omic contain any signal at all?  If STABL
   selects zero features from an omic, that layer may not contribute useful
   information to the integrated analysis.
2. **Baseline** --- the per-omic biomarker lists can be compared to the
   integrated results to see which features survive the joint analysis.

We use `n_bootstraps = 150` throughout this vignette.  Increase to 500-1000 for
publication-quality results.

**Reading the printed summary:** after printing each fit, look at (1) the number
of features selected and (2) the FDP estimate.  A small number of selected
features with a low FDP (< 0.05) is the ideal outcome.

```{r lambda-cytof}
lambda_cytof <- auto_lambda_grid(
  ool_train$x_list$cytof, ool_train$y,
  family = "gaussian", n_lambda = 20
)
```

```{r fit-cytof}
fit_cytof <- stabl_fit(
  x               = ool_train$x_list$cytof,
  y               = ool_train$y,
  lambda_grid     = lambda_cytof,
  family          = "gaussian",
  n_bootstraps    = 150L,
  artificial_type = "random_permutation",
  random_state    = 42L
)
fit_cytof
```

```{r fit-proteomics}
lambda_prot <- auto_lambda_grid(
  ool_train$x_list$proteomics, ool_train$y,
  family = "gaussian", n_lambda = 20
)

fit_prot <- stabl_fit(
  x               = ool_train$x_list$proteomics,
  y               = ool_train$y,
  lambda_grid     = lambda_prot,
  family          = "gaussian",
  n_bootstraps    = 150L,
  artificial_type = "random_permutation",
  random_state    = 42L
)
fit_prot
```

`get_support()` returns the feature names that passed the stability threshold
in each omic.  Compare the two lists: non-overlapping features suggest
omic-specific biomarkers, while any overlap (if the same molecule is measured
in both layers) indicates high-confidence candidates.

```{r per-omic-support}
cat("CyTOF selected features:
")
print(get_support(fit_cytof))

cat("
Proteomics selected features:
")
print(get_support(fit_prot))
```

**Stability path guidance:** each line in the plot is one feature.  Features
with high selection frequencies at moderate penalty levels (the plateau region
of the curve) are the most reliably detected.  Features that spike at very low
penalty (far left) and drop quickly are likely noise rather than signal.

```{r plot-per-omic, fig.alt = "Stability paths for CyTOF and Proteomics"}
plot_stabl_path(fit_cytof,     title = "CyTOF - stability path")
plot_stabl_path(fit_prot,      title = "Proteomics - stability path")
```

---

## 3. Integrated multi-omic workflow

`stabl_multiomic_train_validate()` orchestrates the full pipeline in a single
call.  It accepts training and validation splits and evaluates two complementary
integration strategies:

| Strategy | How it works | Best when |
|---|---|---|
| **Early fusion** | Concatenates all omic matrices into one wide matrix, then runs a *single* STABL fit across all features jointly. | Omics are measured on the same samples; you want a unified biomarker list. |
| **Late fusion** | Fits one STABL + downstream model per omic separately, then learns the optimal weighted combination of predictions on the validation set. | You want to know each omic's independent predictive contribution. |

Key arguments:

| Argument | Purpose |
|---|---|
| `x_train_list` / `x_valid_list` | Named lists of feature matrices; names must match between train and valid. |
| `lambda_grid` | Named list of lambda grids, one per omic (names must match `x_train_list`). |
| `early_fusion` / `late_fusion` | Toggle each strategy on or off independently. |
| `n_iter_lf` | Number of optimisation iterations for the late-fusion weight search. |

The printed `multi_fit` object summarises the result of every strategy and
reports validation metrics (Spearman rho for regression, AUROC for
classification) so you can compare strategies at a glance.

```{r lambda-grid}
# Build per-omic lambda grids
lambda_list <- list(
  cytof      = auto_lambda_grid(ool_train$x_list$cytof,
                                ool_train$y, family = "gaussian", n_lambda = 20),
  proteomics = auto_lambda_grid(ool_train$x_list$proteomics,
                                ool_train$y, family = "gaussian", n_lambda = 20)
)
```

```{r multiomic-fit}
multi_fit <- stabl_multiomic_train_validate(
  x_train_list    = ool_train$x_list,
  y_train         = ool_train$y,
  lambda_grid     = lambda_list,
  x_valid_list    = ool_valid$x_list,
  y_valid         = ool_valid$y,
  family          = "gaussian",
  n_bootstraps    = 150L,
  artificial_type = "random_permutation",
  early_fusion    = TRUE,
  late_fusion     = TRUE,
  n_iter_lf       = 1000L,
  random_state    = 42L
)

multi_fit
```

---

## 4. Exploring results

`multi_fit` is a list with the following elements:

- **`$fits`**: named list of `stabl_fit` objects, one per omic.  These are the
  per-omic STABL results produced *within* the joint pipeline.
- **`$early_fusion`**: result of the early-fusion step (if
  `early_fusion = TRUE`), containing `$fit` (a `stabl_fit` on the concatenated
  matrix).
- **`$late_fusion`**: result of the late-fusion step (if `late_fusion = TRUE`),
  containing `$valid_predictions` (validation predictions) and `$weights`
  (per-omic importance weights).

```{r per-omic-results}
# Per-omic selected features from the joint pipeline
cat("CyTOF (from multi-omic pipeline):
")
print(get_support(multi_fit$fits$cytof))

cat("
Proteomics (from multi-omic pipeline):
")
print(get_support(multi_fit$fits$proteomics))
```

```{r early-fusion-results}
# Early fusion: joint selection across all omics simultaneously
if (!is.null(multi_fit$early_fusion)) {
  cat("Early fusion selected features:\n")
  print(get_support(multi_fit$early_fusion$fit))
}
```

Early fusion features may differ from the per-omic lists because the joint
model sees features from both omics simultaneously.  A feature from one omic
that is redundant with a feature from another omic may drop out of the joint
selection --- this is the expected behaviour and not a bug.

```{r late-fusion-results}
# Late fusion: optimal linear combination of per-omic predictions
if (!is.null(multi_fit$late_fusion)) {
  cat("Late fusion validation predictions (first 6):\n")
  print(head(multi_fit$late_fusion$valid_predictions))
}
```

`$valid_predictions` is a numeric vector of predicted DOS values for the held-out
validation set --- one prediction per sample.  Compare these to the true values
(`ool_valid$y`) to assess predictive performance.  The `$weights` element shows
how much each omic contributed to the combined prediction; a weight near zero
for an omic means it added little beyond the other omic.

```{r stability-paths, fig.alt = "CyTOF stability path from multi-omic fit"}
plot_stabl_path(multi_fit$fits$cytof,
                title = "CyTOF (multi-omic pipeline)")
plot_stabl_path(multi_fit$fits$proteomics,
                title = "Proteomics (multi-omic pipeline)")
```

---

## 5. Export results

`save_stabl_results()` writes four outputs to disk for a single `stabl_fit`
object:

| File | Content |
|---|---|
| `stability_scores.csv` | Feature names and their stability scores. |
| `selected_features.csv` | Only the features above the threshold. |
| `stability_path.pdf` | The stability path plot. |
| `fdr_graph.pdf` | The FDR graph. |

Set `override = TRUE` to overwrite an existing directory; useful in iterative
analyses where you re-run with more bootstraps.

```{r export}
out_dir <- file.path(tempdir(), "stablr_ool_cytof")

save_stabl_results(
  object    = multi_fit$fits$cytof,
  path      = out_dir,
  x         = ool_train$x_list$cytof,
  y         = ool_train$y,
  task_type = "regression",
  override  = TRUE
)

list.files(out_dir, recursive = TRUE)
```

---

## Next steps

- Use `stabl_multiomic_cv()` for a fully cross-validated evaluation when no
  pre-defined validation split is available.
- Pass `groups` to respect subject-level grouping in bootstrap resampling and
  CV fold construction.
- See `vignette("stablr-intro")` for a walk-through of single-omic usage and
  alternative learners (adaptive lasso, elastic net).
