---
title: "Introduction to stablr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to stablr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
set.seed(42)
```

## What is STABL?

STABL (Stability-based Thresholding and Aggregation for Biomarker discovery via
LASSO) is a bootstrap-based method for reliable sparse feature selection in
high-dimensional datasets. It runs many bootstrap resamples of a penalised
learner, computes empirical selection probabilities for each feature, and
controls the false-discovery proportion (FDP) by comparing real features against
artificial decoys.

The `stablr` package provides a pure-R implementation with full
`glmnet`-ecosystem compatibility.  The examples use small simulated data,
random-permutation decoys, and a limited number of bootstraps so that the
vignette remains suitable for routine package builds.

In this setting, a reliable biomarker is not a feature selected once by a
penalised model.  It is a feature selected repeatedly across bootstrap
resamples, with a selection frequency that stands apart from the frequencies
obtained by artificial noise features.

The workflow is:

1. Prepare a numeric feature matrix `x` and an outcome vector `y`.
2. Build a lambda grid with `auto_lambda_grid()`.
3. Fit STABL with `stabl_fit()`.
4. Inspect selected features, stability scores, and diagnostic plots.

The fitted object is a ranked feature-selection result rather than a prediction
model.  It records which features pass the stability threshold, how often each
feature was selected, and how those selections compare with the
artificial-feature reference.

## Installation

```{r install, eval = FALSE}
# From the repository root after cloning:
devtools::install("r-pkg/stablr")
```

```{r load}
library(stablr)
```

## Scope

This vignette covers the single-omic introduction to `stablr`:

1. preparing the input objects expected by `stabl_fit()`;
2. building a compact penalty grid with `auto_lambda_grid()`;
3. fitting a small binary or regression example;
4. extracting selected features (`get_support()`) and stability scores
   (`get_importances()`).
5. reading the stability-path and FDP diagnostic plots;
6. switching between lasso, adaptive lasso, and elastic-net base learners.

## Basic input shape

The examples below use ordinary R objects:

| Object | Expected shape | Notes |
|---|---|---|
| `x` | numeric matrix or matrix-like object, samples in rows and features in columns | Column names become feature names in output. |
| `y` | outcome vector with one value per row of `x` | Use `0`/`1` or a factor-like outcome for `family = "binomial"`; use numeric values for `family = "gaussian"`. |
| `lambda_grid` | data frame returned by `auto_lambda_grid()` | Each row defines a sparse model setting evaluated during bootstrapping. |

If `x` has row names and `y` has names, they should refer to the same samples in
the same order.  This convention is especially important in biomarker analyses,
where a silent sample mismatch can be more damaging than an explicit model
error.

---

## 1. Binary classification

### Simulated data

Binary classification is often the first biomarker-screening setting: rows are
samples, columns are candidate features, and the outcome has two classes.  The
simulated data have 120 samples and 24 features.  Five features (B1-B5) carry a
true signal; the remaining 19 are pure noise.  Because this is a controlled
simulation, the expected number of real features is known before fitting:
`length(signal_features_bin) == 5`.

```{r sim-binary}
n_bin <- 120; p_bin <- 24; n_signal_bin <- 5
signal_features_bin <- paste0("B", seq_len(n_signal_bin))

X_bin <- matrix(rnorm(n_bin * p_bin), nrow = n_bin,
                dimnames = list(paste0("s", seq_len(n_bin)),
                                paste0("B", seq_len(p_bin))))

# True signal in B1-B5; all other features are independent noise.
beta_bin       <- c(rep(1.8, n_signal_bin), rep(0, p_bin - n_signal_bin))
log_odds       <- X_bin %*% beta_bin
y_bin          <- rbinom(n_bin, size = 1, prob = plogis(log_odds))
names(y_bin)   <- rownames(X_bin)
table(y_bin)

signal_features_bin
```

The `table()` output gives the class balance.  This small example is easiest to
interpret when the two classes are not strongly imbalanced; in real analyses,
substantial imbalance should be considered before interpreting the selected
features.

### Fitting STABL

**Step 1 --- build a penalty grid.**  `auto_lambda_grid()` uses `glmnet` to
find the data-dependent range of lambda values.  Every row becomes one lambda
configuration evaluated during bootstrapping.

**Step 2 --- run the bootstrap loop.**  `stabl_fit()` resamples the data
`n_bootstraps` times, fits the penalised learner at each (lambda, alpha) point,
and records which features are selected.  Key arguments:

| Argument | What it does |
|---|---|
| `family` | Passed to `glmnet`; `"binomial"` for 0/1 outcome, `"gaussian"` for continuous. |
| `n_bootstraps` | Number of bootstrap resamples.  The examples use 40-50; final analyses usually need hundreds. |
| `artificial_type` | `"random_permutation"` adds column-permuted decoys; `"knockoff"` uses model-X knockoffs for stronger FDP control. |
| `random_state` | Integer seed for full reproducibility. |

The example uses `n_bootstraps = 50` and keeps the strong-penalty end of a
20-point lambda path.  This gives a short runtime and a compact support.  Lower
Lower penalties admit more features, which can be useful for diagnostics
but can obscure a small planted signal in a short introductory run.  Final analyses
should use more bootstrap resamples and should check that the selected features
are not an artefact of one narrow grid choice.

```{r fit-binary}
lambda_bin_full <- auto_lambda_grid(
  X_bin, y_bin, family = "binomial", n_lambda = 20
)
lambda_bin <- lambda_bin_full[seq_len(6), , drop = FALSE]

fit_bin <- stabl_fit(
  x               = X_bin,
  y               = y_bin,
  lambda_grid     = lambda_bin,
  family          = "binomial",
  n_bootstraps    = 50L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

fit_bin
```

**Reading the printed summary** --- the output reports three key numbers:

- **Stability threshold** --- the minimum selection frequency a feature must
  reach to be called a biomarker (e.g. `0.7` means selected in >= 70% of
  bootstraps).
- **n selected** --- features passing the threshold; for our 5-signal data you
  expect a small number close to 5.
- **FDP+ estimate** --- estimated upper bound on the false-discovery proportion
  among selected features.  With only five planted signals and the additive
  `+1` FDP+ correction, the best possible value for a perfect five-feature
  selection is `1 / 5 = 0.2`; larger real analyses can support smaller values.

In a real analysis, these numbers are best read as a screening summary.  The
most interpretable fits have a selected set of plausible size, a visible gap
between selected and unselected stability scores, and features that remain
credible under expert review.

### Inspecting results

`get_support()` returns a named logical vector for the features that exceed the
stability threshold.  `get_feature_names_out()` returns those names directly.
`get_feature_names_out()` gives the selected biomarker names directly, whereas
`get_support()` is useful when the original feature matrix needs to be subset by
the selected mask.

In this controlled example the planted real features are B1-B5, so the selected
set should match `signal_features_bin`.

`get_importances()` returns a named numeric vector of stability scores
(selection frequencies) for *all* features, sorted from highest to lowest.
A score of 0.8 means the feature was selected in 80% of bootstraps.  The gap
between the top entries and the lower ones shows how cleanly planted features
separate from noise.

```{r results-binary}
# Features selected at the STABL threshold
selected_bin <- get_feature_names_out(fit_bin)
selected_bin

# Compare the selected set against the known simulation truth
setdiff(signal_features_bin, selected_bin)  # missed planted features
setdiff(selected_bin, signal_features_bin)  # selected noise features

# Stability scores for all features (sorted descending)
head(get_importances(fit_bin), 10)
```

The two `setdiff()` calls are only possible because the simulation truth is
known.  In real data, the corresponding evidence comes from the ranked
stability scores, external validation, and biological or clinical review.

### Visualisation

**Stability path** --- each curve traces one feature's selection frequency as
lambda increases (left = weak penalty, many features selected; right = strong
penalty, few features selected).  Features are coloured by whether they exceed
the threshold.  Look for a visible gap between the top-scoring cluster and the
bulk of near-zero lines: that gap is your signal-to-noise contrast.  On the
simulated data B1-B5 should form a distinct upper band.

```{r plot-binary, fig.alt = "Stability path for binary classification"}
plot_stabl_path(fit_bin, title = "Binary classification - stability path")
```

The plotting helpers return ordinary `ggplot` objects, so the same plot can be
displayed in the vignette or saved to disk.

```{r save-path-plot, eval = FALSE}
path_plot <- plot_stabl_path(fit_bin, title = "Binary classification - stability path")
ggplot2::ggsave("stability-path.png", path_plot, width = 7, height = 5)
```

**FDR graph** --- the x-axis sweeps candidate thresholds from 0 to 1; the
y-axis shows the estimated FDP at each threshold.  The dashed vertical line
marks the chosen threshold (minimises FDP while keeping at least one feature);
the horizontal dashed line is the FDP target (default 0.05).  In this tiny
five-signal simulation the FDP+ lower bound is `1 / 5 = 0.2`, so use this graph
as a threshold diagnostic rather than expecting the curve to fall below 0.05.

```{r fdr-binary, fig.alt = "FDR estimate graph for binary classification"}
plot_fdr_graph(fit_bin, title = "Binary classification - FDR estimate")
```

```{r save-fdr-plot, eval = FALSE}
fdr_plot <- plot_fdr_graph(fit_bin, title = "Binary classification - FDR estimate")
ggplot2::ggsave("fdr-graph.png", fdr_plot, width = 7, height = 5)
```

---

## 2. Regression

### Simulated data

The regression workflow differs mainly in the outcome.  With
`family = "gaussian"`, `y` is continuous rather than binary.  This example uses
an independent simulation rather than reusing the binary feature matrix.  Again,
exactly five features (R1-R5) are planted as real before fitting; the remaining
19 are independent noise features.

```{r sim-reg}
n_reg <- 160; p_reg <- 24; n_signal_reg <- 5
signal_features_reg <- paste0("R", seq_len(n_signal_reg))

X_reg <- matrix(rnorm(n_reg * p_reg), nrow = n_reg,
                dimnames = list(paste0("s", seq_len(n_reg)),
                                paste0("R", seq_len(p_reg))))
beta_reg     <- c(rep(3, n_signal_reg), rep(0, p_reg - n_signal_reg))
signal_reg   <- as.numeric(X_reg %*% beta_reg)
y_reg        <- signal_reg + rnorm(n_reg, sd = 0.5)
names(y_reg) <- rownames(X_reg)

signal_features_reg
```

### Fitting STABL

The code mirrors the binary example.  Moving from classification to regression
requires changing `family`, rebuilding the lambda grid for the new outcome, and
refitting the model.

```{r fit-reg}
lambda_reg_full <- auto_lambda_grid(
  X_reg, y_reg, family = "gaussian", n_lambda = 20
)
lambda_reg <- lambda_reg_full[seq_len(6), , drop = FALSE]

fit_reg <- stabl_fit(
  x               = X_reg,
  y               = y_reg,
  lambda_grid     = lambda_reg,
  family          = "gaussian",
  n_bootstraps    = 50L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

fit_reg
```

### Results

The strongest scores should concentrate on R1-R5, the planted regression
features.  You may notice different stability scores compared to the binary fit
--- this is expected because the Gaussian loss function changes the effective
sparsity of each bootstrap model.

```{r results-reg}
selected_reg <- get_feature_names_out(fit_reg)
selected_reg

setdiff(signal_features_reg, selected_reg)  # missed planted features
setdiff(selected_reg, signal_features_reg)  # selected noise features

head(get_importances(fit_reg), 10)
```

The selected feature names give the thresholded screening result.  The full
importance vector keeps the ranked evidence for features just below the chosen
threshold.

### Visualisation

The stability path should look qualitatively similar to the binary case: R1-R5
curve upward and separate from the noise features.  Small differences in curve
heights between the two tasks reflect how each family amplifies feature signal.

```{r plot-reg, fig.alt = "Stability path for regression"}
plot_stabl_path(fit_reg, title = "Regression - stability path")
```

---

## 3. Learner variants

The default base learner is lasso.  The package also exposes variants for
settings in which the selection problem has a different structure.  The examples
below reuse the binary simulation so that the learner choice is the only change.

### Adaptive lasso

The adaptive lasso re-weights the L1 penalty so that features with small
coefficients in a preliminary ridge fit are penalised more heavily.  This
sharpens the stability contrast: true signal features accumulate higher
selection frequencies while noise features are pushed further toward zero.

**When to use it:** prefer adaptive lasso when you expect a sparse true signal
and want a tighter biomarker list with fewer borderline features.

**What to expect:** B1-B5 should remain among the strongest-ranked features,
and `get_importances()` will typically show higher scores for true features
compared to plain lasso.  Borderline noise features can still appear because
this variant chunk uses a broader introductory lambda grid.

```{r adaptive-lasso}
lambda_ada <- auto_lambda_grid(X_bin, y_bin, family = "binomial", n_lambda = 10)

fit_ada <- stabl_fit(
  x               = X_bin,
  y               = y_bin,
  lambda_grid     = lambda_ada,
  base_learner    = "adaptive_lasso",
  family          = "binomial",
  n_bootstraps    = 40L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

get_feature_names_out(fit_ada)
sum(get_support(fit_ada))
```

The count from `sum(get_support(fit_ada))` shows how compact the selected set
has become.  `get_importances(fit_ada)` keeps the full feature ranking.

### Elastic net

Elastic net interpolates between lasso (alpha = 1) and ridge (alpha = 0).
The L2 component encourages correlated features to be selected together, which
is useful when biomarkers form functional modules (e.g. co-expressed genes).

Passing multiple `l1_ratio` values to `auto_lambda_grid()` makes STABL sweep
the (lambda, alpha) grid jointly, exploring both the sparsity level and the
L1/L2 mix in every bootstrap.

**When to use it:** prefer elastic net over plain lasso when features are highly
correlated and you want to avoid arbitrarily picking one representative from a
correlated group.

```{r elastic-net}
lambda_en <- auto_lambda_grid(
  X_bin, y_bin,
  family = "binomial", n_lambda = 8,
  l1_ratio = c(0.5, 0.9)
)

fit_en <- stabl_fit(
  x               = X_bin,
  y               = y_bin,
  lambda_grid     = lambda_en,
  base_learner    = "elastic_net",
  family          = "binomial",
  n_bootstraps    = 40L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

get_feature_names_out(fit_en)
```

---

## Next steps

This vignette is limited to the single-omic package basics: fitting a model,
inspecting selected features, viewing diagnostics, and trying common learner
variants.  Further entry points are:

- Use more bootstraps for final analyses, typically hundreds rather than the
  small values used here.
- Use `vignette("stablr-multiomic")` when your data contain multiple omics
  views.
- Use `?save_stabl_results` when you are ready to export scores, plots, and
  selected-feature tables to disk.
