---
title: "Python-R Workflow Mapping for STABL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Python-R Workflow Mapping for STABL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  cache      = TRUE,
  cache.path = "stablr-python-parity-cache/"
)
```

## Overview

This vignette maps the analyses from the STABL Tutorial Notebook onto the R
`stablr` API.  It uses the same preprocessing shape and real OOL validation
workflow, but the rendered run uses bounded settings so package builds remain
practical.  Use the larger settings noted below for a high-fidelity parity
experiment.

| Aspect | Python (`stabl`) | R (`stablr`) |
|---|---|---|
| Base learner | `sklearn.linear_model.Lasso` | `glmnet::glmnet` (LASSO) |
| Bootstrapping | `n_bootstraps` | `n_bootstraps` |
| FDP control | `fdr_threshold_range = np.arange(0, 1, 0.01)` | `seq(0, 1, by = 0.01)` |
| Artificial features | `"knockoff"` in the tutorial | `random_permutation` here for render speed; `knockoff` for extended parity |
| Random state | `42` | `random_state = 42L` |

**Two analyses:**

1. **OOL Regression** -- Onset of Labor proteomics, `family = "gaussian"`,
   real train/validation split, bounded STABL settings.
2. **COVID-19 Binary Classification** -- Proteomics, `family = "binomial"`,
   shown as an extended optional run because the data are not bundled with the
   package build.

```{r libraries}
library(stablr)

render_n_bootstraps <- 120L
render_n_lambda <- 8L
render_artificial_type <- "random_permutation"
run_extended <- identical(Sys.getenv("STABLR_RUN_EXTENDED_VIGNETTES"), "true")
```

The rendered configuration uses `r render_n_bootstraps` bootstraps and
`r render_n_lambda` lambda values.  For a parity report intended to match the
Python tutorial more closely, use 500-1000 bootstraps with
`artificial_type = "knockoff"` and run outside package-build time.

---

## Preprocessing helpers

The Python tutorial uses: `VarianceThreshold -> LowInfoFilter -> SimpleImputer (median) -> StandardScaler`.

```{r preprocess-helpers}
preprocess_fit <- function(x, min_var = 1e-6) {
  col_vars  <- apply(x, 2, var, na.rm = TRUE)
  keep_cols <- col_vars > min_var
  x <- x[, keep_cols, drop = FALSE]
  col_medians <- apply(x, 2, median, na.rm = TRUE)
  for (j in seq_along(col_medians)) {
    na_idx <- is.na(x[, j])
    if (any(na_idx)) x[na_idx, j] <- col_medians[j]
  }
  col_means <- colMeans(x)
  col_sds   <- apply(x, 2, sd)
  col_sds[col_sds < 1e-10] <- 1
  x <- sweep(sweep(x, 2, col_means), 2, col_sds, "/")
  list(x = x, keep_cols = keep_cols, col_medians = col_medians,
       col_means = col_means, col_sds = col_sds)
}

preprocess_apply <- function(x, pipe) {
  x <- x[, pipe$keep_cols, drop = FALSE]
  for (j in seq_len(ncol(x))) {
    na_idx <- is.na(x[, j])
    if (any(na_idx)) x[na_idx, j] <- pipe$col_medians[j]
  }
  sweep(sweep(x, 2, pipe$col_means), 2, pipe$col_sds, "/")
}
```

---

## Part 1 -- Regression: Onset of Labor Proteomics

```{r ool-load}
ool_train <- load_ool_data(split = "train")
ool_valid <- load_ool_data(split = "valid")

pipe_prot    <- preprocess_fit(ool_train$x_list$proteomics)
x_prot       <- pipe_prot$x
y_prot       <- ool_train$y
x_prot_valid <- preprocess_apply(ool_valid$x_list$proteomics, pipe_prot)
y_prot_valid <- ool_valid$y

cat("Training set:   ", nrow(x_prot), "samples x", ncol(x_prot), "features\n")
cat("Validation set: ", nrow(x_prot_valid), "samples x", ncol(x_prot_valid), "features\n")
```

```{r ool-lambda}
lambda_prot <- auto_lambda_grid(
  x_prot, y_prot,
  family   = "gaussian",
  n_lambda = render_n_lambda
)
```

```{r ool-fit, cache = TRUE}
set.seed(42)
fit_prot <- stabl_fit(
  x                   = x_prot,
  y                   = y_prot,
  lambda_grid         = lambda_prot,
  base_learner        = "lasso",
  family              = "gaussian",
  n_bootstraps        = render_n_bootstraps,
  artificial_type     = render_artificial_type,
  fdr_threshold_range = seq(0, 1, by = 0.01),
  random_state        = 42L
)
fit_prot
```

```{r ool-results}
sel_prot <- get_support(fit_prot)
cat("Selected features (n =", sum(sel_prot), "):\n")
print(names(which(sel_prot)))
cat("\nTop 10 stability scores:\n")
print(head(sort(get_importances(fit_prot), decreasing = TRUE), 10))
```

```{r ool-fdr}
plot_fdr_graph(fit_prot, title = "OOL Proteomics -- FDP+ curve (regression)")
```

```{r ool-path}
plot_stabl_path(fit_prot, title = "OOL Proteomics -- stability path (regression)")
```

```{r ool-scatter, fig.height = 6}
sel_prot_names <- names(which(sel_prot))
if (length(sel_prot_names) > 0) {
  scatterplot_features(
    features = sel_prot_names,
    x        = x_prot,
    y        = y_prot,
    title    = "OOL Proteomics -- selected features vs DOS"
  )
}
```

---

## Part 2 -- Binary Classification: COVID-19 Proteomics

```{r covid-check}
covid_train_dir <- normalizePath(
  file.path(system.file("", package = "stablr"),
            "..", "..", "..", "..", "Sample Data", "COVID-19", "Training"),
  mustWork = FALSE
)
eval_covid <- dir.exists(covid_train_dir) &&
              file.exists(file.path(covid_train_dir, "Proteomics.csv")) &&
              isTRUE(run_extended)
if (!dir.exists(covid_train_dir)) {
  message("COVID-19 data not found -- skipping COVID-19 section.")
} else if (!isTRUE(run_extended)) {
  message("COVID-19 data found, but the section is skipped by default. Set STABLR_RUN_EXTENDED_VIGNETTES=true to evaluate it.")
}
```

```{r covid-load, eval = eval_covid}
covid_valid_dir <- file.path(dirname(covid_train_dir), "Validation")

prot_tr  <- as.matrix(read.csv(file.path(covid_train_dir, "Proteomics.csv"),
                                row.names = 1, check.names = FALSE))
y_tr_df  <- read.csv(file.path(covid_train_dir, "Mild&ModVsSevere.csv"),
                      row.names = 1, check.names = FALSE)
prot_val <- as.matrix(read.csv(file.path(covid_valid_dir, "Validation_proteomics.csv"),
                                row.names = 1, check.names = FALSE))
y_val_df <- read.csv(
  file.path(covid_valid_dir, "Validation_outcome(WHO.0 >= 5).csv"),
  row.names = 1, check.names = FALSE)

common_tr  <- intersect(rownames(prot_tr),  rownames(y_tr_df))
common_val <- intersect(rownames(prot_val), rownames(y_val_df))

pipe_covid    <- preprocess_fit(prot_tr[common_tr, ])
x_covid_train <- pipe_covid$x
y_covid_train <- setNames(as.integer(y_tr_df[common_tr, 1]), common_tr)
x_covid_valid <- preprocess_apply(prot_val[common_val, ], pipe_covid)
y_covid_valid <- setNames(as.integer(y_val_df[common_val, 1]), common_val)

cat("COVID train:", nrow(x_covid_train), "x", ncol(x_covid_train), "\n")
cat("COVID valid:", nrow(x_covid_valid), "x", ncol(x_covid_valid), "\n")
print(table(y_covid_train))
```

```{r covid-lambda, eval = eval_covid}
lambda_covid <- auto_lambda_grid(
  x_covid_train, y_covid_train,
  family   = "binomial",
  n_lambda = render_n_lambda
)
```

```{r covid-fit, cache = TRUE, eval = eval_covid}
set.seed(42)
fit_covid <- stabl_fit(
  x                   = x_covid_train,
  y                   = y_covid_train,
  lambda_grid         = lambda_covid,
  base_learner        = "lasso",
  family              = "binomial",
  n_bootstraps        = 200L,
  artificial_type     = render_artificial_type,
  fdr_threshold_range = seq(0.1, 1, by = 0.01),
  random_state        = 42L
)
fit_covid
```

```{r covid-results, eval = eval_covid}
sel_covid <- get_support(fit_covid)
cat("Selected features (n =", sum(sel_covid), "):\n")
print(names(which(sel_covid)))
```

```{r covid-fdr, eval = eval_covid}
plot_fdr_graph(fit_covid, title = "COVID-19 Proteomics -- FDP+ curve")
```

```{r covid-path, eval = eval_covid}
plot_stabl_path(fit_covid, title = "COVID-19 Proteomics -- stability path")
```

```{r covid-boxplot, fig.height = 6, eval = eval_covid}
sel_covid_names <- names(which(sel_covid))
if (length(sel_covid_names) > 0) {
  boxplot_features(
    features = sel_covid_names,
    x        = x_covid_train,
    y        = factor(y_covid_train, labels = c("Mild/Mod", "Severe")),
    title    = "COVID-19 -- selected features (training)"
  )
}
```

---

## Part 3 -- Full pipeline: validation performance

```{r ool-pipeline}
sel_names_prot <- names(which(get_support(fit_prot)))
cat("OOL: using", length(sel_names_prot), "selected features for final model\n")

if (length(sel_names_prot) > 0) {
  df_tr <- as.data.frame(x_prot[, sel_names_prot, drop = FALSE])
  df_tr$y <- y_prot
  final_lm <- lm(y ~ ., data = df_tr)

  df_val <- as.data.frame(x_prot_valid[, sel_names_prot, drop = FALSE])
  y_pred_val <- predict(final_lm, newdata = df_val)

  cat("Validation Pearson r:", round(cor(y_pred_val, y_prot_valid), 3), "\n")
  cat("Validation RMSE:     ", round(sqrt(mean((y_pred_val - y_prot_valid)^2)), 2), "\n")
} else {
  cat("No features selected -- cannot build final model.\n")
}
```

```{r covid-pipeline, eval = eval_covid}
sel_names_covid <- names(which(get_support(fit_covid)))
cat("COVID-19: using", length(sel_names_covid), "selected features for final model\n")

if (length(sel_names_covid) > 0) {
  df_tr_c <- as.data.frame(x_covid_train[, sel_names_covid, drop = FALSE])
  df_tr_c$y <- y_covid_train
  final_glm <- glm(y ~ ., data = df_tr_c, family = binomial())

  df_val_c <- as.data.frame(x_covid_valid[, sel_names_covid, drop = FALSE])
  y_pred_covid <- predict(final_glm, newdata = df_val_c, type = "response")

  plot_roc(y_true = y_covid_valid, y_preds = y_pred_covid,
           title = "COVID-19 -- Validation ROC")
  plot_prc(y_true = y_covid_valid, y_preds = y_pred_covid,
           title = "COVID-19 -- Validation PRC")
} else {
  cat("No features selected -- cannot build final model.\n")
}
```

---

## Parity notes

- `glmnet` LASSO and scikit-learn LASSO differ in coordinate descent
  implementation. Feature selections should be concordant but not identical.
- The rendered vignette uses random-permutation decoys for speed.  For
  high-fidelity parity work, switch to `artificial_type = "knockoff"` and use
  the Python tutorial bootstrap counts.
- Stability scores should be compared at matched bootstrap counts, artificial
  feature strategy, lambda grid, and preprocessing choices.

```{r session-info}
sessionInfo()
```
