---
title: "Cooperative Fusion for Multi-Omic Biomarker Discovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cooperative Fusion for Multi-Omic Biomarker Discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
set.seed(42)
```

```{r multiview-check, include = FALSE}
# Cooperative fusion requires the optional multiview package.
# If it is not installed, code chunks are shown but not evaluated.
if (!requireNamespace("multiview", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
  message(
    "The 'multiview' package is not installed. ",
    "Code chunks will be displayed but not evaluated. ",
    "Install it with: install.packages('multiview') ",
    "or R CMD INSTALL <path/to/multiview>."
  )
}
```

## Overview

This vignette demonstrates **cooperative fusion** --- a third multi-omic
integration strategy in `stablr` that sits between early and late fusion.
Cooperative fusion fits the views *jointly* via the `multiview` package while
letting each view retain its own coefficient path, allowing the omics to
"cooperate" on shared signal rather than competing in a concatenated space.

**When should you use cooperative fusion?**

| Strategy | How it works | Best when |
|---|---|---|
| **Per-omic** | One STABL fit per omic, independently. | Baseline; or when views are likely orthogonal. |
| **Early fusion** | Concatenate all omics, one joint STABL fit. | Omics are tightly correlated; you want a single feature list. |
| **Late fusion** | Per-omic fits, then learn a weighted blend of predictions. | Each omic may predict independently; you want omic weights. |
| **Cooperative fusion** | Joint multiview fit with cooperation strength `rho`; each view has its own coefficient path. | Omics share some signal structure; you want cross-view feature selection to be coherent. |

The cooperation strength parameter `rho` controls how much views are encouraged
to share their signal.  At `rho = 0` cooperative fusion reduces to independent
fits; as `rho` grows, views penalise each other towards a common prediction.
`stablr` tunes `rho` automatically over a user-supplied grid.

## Learning objectives

By the end of this vignette you will be able to:

1. Explain what `rho` controls in cooperative fusion and how to choose a grid.
2. Run `stabl_multiomic_train_validate()` with `cooperative_fusion = TRUE` and
   navigate the `$cooperative_fusion` result element.
3. Read the `print()` output for cooperative fusion details.
4. Understand which `cooperation_selection` and `cooperation_selector`
   combinations are supported and why.
5. Compare cooperative fusion feature selections to early and late fusion on the
   same dataset.
6. Use `stabl_multiomic_cv()` to evaluate cooperative fusion across outer folds.

---

## 1. Load the data

We use the same Onset of Labor (OOL) dataset as `vignette("stablr-multiomic")`:
two omic layers (CyTOF immune profiling + Proteomics) in 150 pregnant women,
with days-to-onset-of-labor (DOS) as the continuous regression target.

```{r load}
library(stablr)
```

```{r load-data}
ool_train <- load_ool_data(split = "train")
ool_valid <- load_ool_data(split = "valid")

# Shape of each omic layer
lapply(ool_train$x_list, dim)
```

---

## 2. Build lambda grids

Cooperative fusion requires a lambda grid for each omic, exactly as in
`vignette("stablr-multiomic")`.  We build one grid per omic with
`auto_lambda_grid()` and pass them as a named list.

```{r lambda-grids}
lambda_list <- list(
  cytof      = auto_lambda_grid(
    ool_train$x_list$cytof, ool_train$y,
    family = "gaussian", n_lambda = 20
  ),
  proteomics = auto_lambda_grid(
    ool_train$x_list$proteomics, ool_train$y,
    family = "gaussian", n_lambda = 20
  )
)

# Number of lambda values per omic
vapply(lambda_list, nrow, integer(1L))
```

---

## 3. Running cooperative fusion

Activate cooperative fusion by setting `cooperative_fusion = TRUE` and
supplying a `rho` grid.  `stablr` fits one multiview model per `rho` value
(using shared inner CV folds), then picks the best `rho` by the chosen
`cooperation_type_measure`.

Key arguments:

| Argument | Default | Purpose |
|---|---|---|
| `rho` | `0` | Scalar or vector of cooperation strengths to try. |
| `cooperation_selection` | `"cv"` | `"cv"` (inner CV) or `"validation"` (held-out set). |
| `cooperation_selector` | `"lambda.min"` | Which lambda to pick: `"lambda.min"` or `"lambda.1se"` (CV only). |
| `cooperation_type_measure` | `"default"` | Tuning metric; defaults to `"mse"` for Gaussian. |
| `cooperation_nfolds` | `5L` | Number of inner CV folds when `cooperation_selection = "cv"`. |

```{r cooperative-fit}
cf_fit <- stabl_multiomic_train_validate(
  x_train_list          = ool_train$x_list,
  y_train               = ool_train$y,
  lambda_grid           = lambda_list,
  x_valid_list          = ool_valid$x_list,
  y_valid               = ool_valid$y,
  family                = "gaussian",
  n_bootstraps          = 50L,
  artificial_type       = "random_permutation",
  random_state          = 42L,
  cooperative_fusion    = TRUE,
  rho                   = c(0, 0.1, 0.3, 0.5),
  cooperation_selection = "cv",
  cooperation_selector  = "lambda.min",
  cooperation_nfolds    = 5L
)
```

`print()` shows a summary of every active branch.  When cooperative fusion is
present you will see a `Cooperative fusion:` block at the bottom:

```{r print-fit}
cf_fit
```

**Reading the cooperative fusion block:**

- **`selection`** --- how `rho` was tuned: `"cv"` here.
- **`rho (chosen)`** --- the `rho` value selected by cross-validation.  A value
  of `0` means the data did not benefit from cross-view cooperation; a larger
  value means views actively helped each other.
- **`selector`** --- which lambda rule was applied (`lambda.min`).
- **`type.measure`** --- the CV metric used for tuning (e.g., `mse`).
- **`score`** --- the CV score at the chosen `rho` and `lambda`.
- **`selected feats`** --- total number of features selected across all views.

---

## 4. Navigating `$cooperative_fusion`

The `cooperative_fusion` element of the result list contains detailed tuning
output.

```{r cf-element}
cf <- cf_fit$cooperative_fusion

# Chosen hyperparameters
cat("Chosen rho:     ", cf$rho, "\n")
cat("Chosen lambda:  ", round(cf$selected_lambda, 4), "\n")
cat("Selector:       ", cf$selector, "\n")
cat("CV score (MSE): ", round(cf$score, 4), "\n")
```

### Selected features per view

`$selected_features` is a named list with one character vector per omic:

```{r cf-selected}
cat("CyTOF features selected by cooperative fusion:\n")
print(cf$selected_features$cytof)

cat("\nProteomics features selected by cooperative fusion:\n")
print(cf$selected_features$proteomics)
```

### Predictions

`$train_predictions` and `$valid_predictions` hold the multiview predictions
from the chosen model:

```{r cf-predictions}
cat("First 6 training predictions:\n")
print(head(cf$train_predictions))

cat("\nFirst 6 validation predictions:\n")
print(head(cf$valid_predictions))
```

### Tuning diagnostics

`$diagnostics` records the CV score at every `rho` in the grid, making it
possible to inspect the tuning surface:

```{r cf-diagnostics}
# One row per rho value
print(cf$diagnostics[, c("rho", "lambda", "metric_value", "selected")])
```

The `selected` column marks the winning `rho`.  If the surface is flat (similar
`metric_value` across `rho`), the omics do not benefit strongly from
cooperation and `rho = 0` is a safe choice.

---

## 5. Policy constraints

`stablr` enforces two hard constraints to prevent silently incorrect analyses:

### Constraint 1 --- `lambda.1se` requires CV mode

`lambda.1se` applies the "one-standard-error" rule and is only well-defined
when cross-validation is used to estimate the error surface.  Requesting it
with `cooperation_selection = "validation"` raises an error:

```{r guard-lambda1se, error = TRUE}
stabl_multiomic_train_validate(
  x_train_list          = ool_train$x_list,
  y_train               = ool_train$y,
  lambda_grid           = lambda_list,
  x_valid_list          = ool_valid$x_list,
  y_valid               = ool_valid$y,
  family                = "gaussian",
  n_bootstraps          = 3L,
  artificial_type       = NULL,
  cooperative_fusion    = TRUE,
  rho                   = 0,
  cooperation_selection = "validation",
  cooperation_selector  = "lambda.1se"
)
```

### Constraint 2 --- Cox family cannot use validation-based tuning

The multiview Cox path does not support external validation scoring, so
`cooperation_selection = "validation"` with `family = "cox"` is not supported.
Use `cooperation_selection = "cv"` for Cox models:

```{r guard-cox-validation, error = TRUE}
library(survival)

set.seed(42)
n <- 20L
ids <- paste0("s", seq_len(n))
x_a <- matrix(rnorm(n * 4), nrow = n,
               dimnames = list(ids, paste0("a", seq_len(4))))
x_b <- matrix(rnorm(n * 4), nrow = n,
               dimnames = list(ids, paste0("b", seq_len(4))))
y_surv <- Surv(time = rexp(n), event = rbinom(n, 1L, 0.7))
rownames(y_surv) <- ids

stabl_multiomic_train_validate(
  x_train_list          = list(omic_a = x_a, omic_b = x_b),
  y_train               = y_surv,
  lambda_grid           = data.frame(lambda = c(0.2, 0.1)),
  x_valid_list          = list(omic_a = x_a, omic_b = x_b),
  y_valid               = y_surv,
  family                = "cox",
  n_bootstraps          = 2L,
  artificial_type       = NULL,
  cooperative_fusion    = TRUE,
  rho                   = 0,
  cooperation_selection = "validation"
)
```

---

## 6. What happens when `multiview` is not installed

Cooperative fusion is guarded by `requireNamespace("multiview")`.  If
`multiview` is absent and you set `cooperative_fusion = TRUE`, `stablr` stops
immediately with an informative error before any computation runs:

```
Error: `cooperative_fusion = TRUE` requires the optional 'multiview' package
to be installed.
```

Install the bundled version with:

```r
# From the stablr repository root:
R CMD INSTALL multiview
# or from CRAN when available:
install.packages("multiview")
```

Setting `cooperative_fusion = FALSE` (the default) never triggers this check,
so all other `stablr` functionality works without `multiview`.

---

## 7. Comparing cooperative, early, and late fusion

Run all three strategies in a single call and compare which features each
selects:

```{r all-strategies}
all_fit <- stabl_multiomic_train_validate(
  x_train_list          = ool_train$x_list,
  y_train               = ool_train$y,
  lambda_grid           = lambda_list,
  x_valid_list          = ool_valid$x_list,
  y_valid               = ool_valid$y,
  family                = "gaussian",
  n_bootstraps          = 50L,
  artificial_type       = "random_permutation",
  random_state          = 42L,
  early_fusion          = TRUE,
  late_fusion           = TRUE,
  n_iter_lf             = 500L,
  cooperative_fusion    = TRUE,
  rho                   = c(0, 0.1, 0.3, 0.5),
  cooperation_selection = "cv",
  cooperation_selector  = "lambda.min",
  cooperation_nfolds    = 5L
)
```

```{r compare-print}
# print() now shows all four branches
all_fit
```

Compare feature sets across strategies:

```{r compare-features}
ef_features  <- all_fit$early_fusion$selected_features
cf_features  <- unique(unlist(all_fit$cooperative_fusion$selected_features,
                              use.names = FALSE))
per_omic_features <- unique(unlist(all_fit$selected_features,
                                   use.names = FALSE))

cat("Early fusion selected:      ", length(ef_features), "features\n")
cat("Cooperative fusion selected:", length(cf_features), "features\n")
cat("Per-omic union selected:    ", length(per_omic_features), "features\n")

cat("\nIn cooperative but not early fusion:\n")
print(setdiff(cf_features, ef_features))

cat("\nIn early fusion but not cooperative:\n")
print(setdiff(ef_features, cf_features))
```

**Interpreting differences:** Early fusion sees all features simultaneously in
a single penalised model; cooperative fusion fits per-view paths that are
encouraged to agree.  Features unique to cooperative fusion may carry cross-view
signal that is diluted in the concatenated early-fusion space.

---

## 8. Cross-validation with cooperative fusion

Use `stabl_multiomic_cv()` when no pre-defined validation split is available.
Cooperative fusion parameters propagate to every outer fold automatically:

```{r cv-fit}
cv_fit <- stabl_multiomic_cv(
  x_list                = ool_train$x_list,
  y                     = ool_train$y,
  lambda_grid           = lambda_list,
  v                     = 3L,
  family                = "gaussian",
  n_bootstraps          = 20L,
  artificial_type       = "random_permutation",
  random_state          = 42L,
  cooperative_fusion    = TRUE,
  rho                   = c(0, 0.3),
  cooperation_selection = "cv",
  cooperation_selector  = "lambda.min",
  cooperation_nfolds    = 3L
)

cv_fit
```

The `$diagnostics` data frame includes cooperative-specific columns
(`cooperative_rho`, `cooperative_lambda`, `cooperative_score`, etc.) so you can
inspect how `rho` was selected in each outer fold:

```{r cv-diagnostics}
print(cv_fit$diagnostics[, c("fold", "omic", "n_selected",
                              "cooperative_rho", "cooperative_score")])
```

---

## 9. Choosing `rho` and design guidance

| Decision | Guidance |
|---|---|
| **Grid density** | Start with `c(0, 0.1, 0.3, 0.5)`.  Add finer points only if the CV surface has a clear non-zero minimum. |
| **Grid upper bound** | Values > 0.5 can over-penalise disagreement between views; rarely beneficial unless views are near-identical. |
| **`lambda.min` vs `lambda.1se`** | `lambda.1se` gives a sparser model at the cost of a higher CV error; prefer `lambda.min` when prediction is the goal. |
| **`cooperation_nfolds`** | 5 (default) is suitable for n >= 50.  Use 3 for smaller datasets.  Must be >= 3. |
| **`cooperation_selection = "validation"`** | Use when you have a true held-out set and want to tune `rho` and `lambda` together on it.  Not supported for `family = "cox"`. |
| **rho = 0 wins** | If the chosen `rho` is always 0, the views are not benefiting from cooperation.  Consider whether the omics share any biological pathway. |

---

## Next steps

- See `vignette("stablr-multiomic")` for a walk-through of per-omic, early,
  and late fusion on the same OOL dataset.
- See `vignette("stablr-intro")` for single-omic STABL and learner options.
- Pass `groups` to `stabl_multiomic_train_validate()` to prevent leakage in
  longitudinal designs; the grouped bootstrap and outer-fold construction both
  respect this argument.
- For binomial and Cox outcomes, set `family = "binomial"` or `family = "cox"`
  and adjust `cooperation_type_measure` (e.g., `"auc"` for binomial,
  `"deviance"` for Cox).
