---
title: "TCGA Breast Cancer Nested-CV Head-to-Head: stablr vs DIABLO"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCGA Breast Cancer Nested-CV Head-to-Head: stablr vs DIABLO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4,
  cache      = FALSE,
  # eval=FALSE: vignette requires large TCGA data and nested-CV fits;
  # gate from R CMD check and CI. Run interactively or via pkgdown::build_site().
  eval       = FALSE
)
```

## Overview

The compact TCGA vignette shows the workflow.  This vignette asks for evidence.
It compares `stablr` and mixOmics DIABLO on the same three-class PAM50 breast
cancer task: Basal, Her2, and LumA.

The design is nested cross-validation.  Outer folds estimate generalisation.
Inner folds tune the models.  Feature selection is repeated inside each
training split, rather than performed once on the full dataset before
validation.  That distinction is the difference between a benchmark and an
optimistic story.

The head-to-head uses mRNA and miRNA only.  The protein block in
`mixOmics::breast.TCGA` is available only for the original training samples,
while all 220 samples have mRNA and miRNA.  Excluding protein keeps both
methods on the same samples and blocks.

The heavy computation is not run while rendering this vignette.  It is executed
through SLURM and cached as an RDS file.

## Methodological Audit

This workflow was checked against the `nestedcv` vignette guidance for p >> n
biomedical prediction.  No filter or feature-selection result is computed on
the full dataset before cross-validation.  STABL selection, downstream
multinomial refitting, DIABLO component tuning, and DIABLO `keepX` tuning all
occur inside the relevant training split before a held-out outer fold is
predicted.

The design also follows p >> n variable-selection practice:

- predictors are passed as numeric matrices and model scaling is left inside the
  fitting procedures rather than learned globally before CV;
- the outer loop estimates performance, while the inner loop selects candidates
  and tuning parameters;
- stablr uses knockoff artificial features for FDP-aware stability selection;
- the deliverable is a recurrence profile over outer folds, not a single
  definitive biomarker list;
- no naive post-selection p-values are reported.

The benchmark is still exploratory.  Confirmatory extensions would include
external cohort validation and refitting unpenalised subtype models on only the
most recurrent features, with explicit post-selection qualification.

## Run on SLURM

```{r cache-setup}
analysis_script <- file.path("..", "inst", "analysis", "run_tcga_nestedcv.R")
if (!file.exists(analysis_script)) {
  analysis_script <- file.path("r-pkg", "stablr", "inst", "analysis", "run_tcga_nestedcv.R")
}
source(analysis_script)

cache_path <- file.path("..", "inst", "analysis", "cache", "tcga_nestedcv_results.rds")
if (!file.exists(cache_path)) {
  cache_path <- file.path("r-pkg", "stablr", "inst", "analysis", "cache", "tcga_nestedcv_results.rds")
}

slurm_cmd <- "sbatch r-pkg/stablr/inst/analysis/tcga_nestedcv.slurm"
cat("Expected cache:", cache_path, "\n")
cat("Submit command:", slurm_cmd, "\n")
```

```{r cache-status}
has_results <- tcga_cache_complete(cache_path)
if (!has_results) {
  cat(
    "Nested-CV cache is not available yet.\n",
    "Submit the SLURM job above, then render this notebook again.\n",
    sep = ""
  )
}
```

## Analysis Design

The cache records the exact settings used by the SLURM runner:

```{r design-summary, eval = has_results}
res <- readRDS(cache_path)
str(res$settings)
```

`stablr` uses multinomial stability selection with knockoff artificial
features.  The candidate workflows are mRNA-only, miRNA-only, and early fusion
across mRNA and miRNA.  Candidate choice is based on inner-CV balanced error
rate.

DIABLO uses the mixOmics multiblock sPLS-DA workflow: `block.plsda()` and
`perf()` choose the number of components, `tune.block.splsda()` tunes `keepX`,
and the locked outer-fold model is evaluated with weighted-vote
`centroids.dist` predictions.

## Predictive Performance

The performance table is the first readout.  The confusion matrices show where
the methods make their mistakes, which is often more informative than a single
accuracy number.

```{r performance, eval = has_results}
res$performance
```

```{r confusion, eval = has_results}
stabl_conf <- table(
  Truth = res$stablr$outer_predictions$truth,
  Predicted = res$stablr$outer_predictions$predicted
)
diablo_conf <- table(
  Truth = res$diablo$outer_predictions$truth,
  Predicted = res$diablo$outer_predictions$predicted
)

cat("stablr confusion matrix:\n")
print(stabl_conf)
cat("\nDIABLO confusion matrix:\n")
print(diablo_conf)
```

```{r paired-fold-performance, eval = has_results}
fold_metric <- function(pred) {
  by_fold <- split(pred, pred$fold_id)
  do.call(rbind, lapply(names(by_fold), function(id) {
    m <- tcga_metrics(by_fold[[id]]$truth, by_fold[[id]]$predicted)
    data.frame(fold_id = id, m, stringsAsFactors = FALSE)
  }))
}

stabl_fold <- fold_metric(res$stablr$outer_predictions)
diablo_fold <- fold_metric(res$diablo$outer_predictions)
paired <- merge(stabl_fold, diablo_fold, by = "fold_id",
                suffixes = c("_stablr", "_DIABLO"))
paired$delta_BER <- paired$balanced_error_rate_stablr -
  paired$balanced_error_rate_DIABLO
paired$delta_accuracy <- paired$accuracy_stablr - paired$accuracy_DIABLO
summary(paired[, c("delta_BER", "delta_accuracy")])
```

## Feature Comparison

Feature recurrence is the biomarker-side counterpart to predictive
performance.  A recurrent feature is not merely selected once; it survives
across many outer training splits.

```{r feature-overlap, eval = has_results}
res$feature_comparison$overlap
```

```{r feature-recurrence, eval = has_results}
recurrence <- res$feature_comparison$recurrence
head(recurrence[order(recurrence$block, -recurrence$recurrence), ], 30)
```

```{r shared-features, eval = has_results}
shared <- do.call(rbind, lapply(split(recurrence, recurrence$block), function(tab) {
  stabl_feats <- tab$feature[tab$method == "stablr"]
  diablo_feats <- tab$feature[tab$method == "DIABLO"]
  tab[tab$feature %in% intersect(stabl_feats, diablo_feats), ]
}))
shared[order(shared$block, shared$feature, shared$method), ]
```

## Exported Artifacts

The SLURM runner writes these files next to the cache:

```{r exported-files, eval = has_results}
list.files(dirname(cache_path), pattern = "tcga_nestedcv", full.names = TRUE)
```

The CSV files are intended for manuscript tables and supplement material:
performance, feature recurrence, and feature overlap by omic block.

## Interpreting stablr vs DIABLO results

Read predictive and feature-level outputs together.  A method that wins on
balanced error rate (BER) is not automatically preferable for biomarker
discovery, and vice versa.

### Predictive layer

| Observation | Interpretation |
|-------------|----------------|
| Lower outer-fold BER for DIABLO | Latent multiblock structure may explain class separation better than sparse per-feature STABL + refit on this cohort. |
| Lower outer-fold BER for stablr | Inner-CV candidate choice (mRNA / miRNA / early fusion) plus multinomial refit may capture signal DIABLO's fixed component budget misses. |
| Similar BER, different confusion patterns | Methods trade errors across classes; inspect per-class recall before declaring a winner. |
| Large fold-to-fold BER variance | Dataset is small (220 samples); treat point estimates as exploratory. |

Negative `delta_BER` (stablr − DIABLO) favours DIABLO; positive favours stablr.
Report paired fold deltas with bootstrap confidence intervals (see
`inst/analysis/BENCHMARK_PLAN.md`, Phase 2).

### Feature / biomarker layer

| Observation | Interpretation |
|-------------|----------------|
| High recurrence, modest overlap with DIABLO | Stability selection and component sparsity surface partially complementary feature sets. |
| stablr features recurrent, absent from DIABLO loadings | May reflect block-specific marginal signal stabilised by resampling, not captured in shared latent factors. |
| DIABLO `selectVar` features absent from stablr support | May load strongly on components without surviving FDP+ thresholding across bootstraps. |
| stablr selects more features per fold | FDP threshold and lambda grid control sparsity; DIABLO sparsity is governed by `keepX`. |

### When each method is the better tool

| Criterion | Prefer stablr | Prefer DIABLO |
|-----------|---------------|---------------|
| Primary deliverable | Named biomarker list with stability evidence | Integrative classification + component visualisation |
| False-discovery control | Knockoff / FDP+ built in | Not part of the default workflow |
| Outcome types | gaussian, binomial, multinomial, Cox | Classification-centric multiblock sPLS-DA |
| Block coupling | Cooperative `rho` or fusion modes | Design matrix + shared latent space |
| Ecosystem | glmnet, native R, Python STABL parity | mixOmics tutorials, circos, network plots |
| Compute profile | Many bootstraps × lambda grid | Component + `keepX` tuning (typically faster per fold) |

For Python STABL parity (selection semantics, not prediction contests), see
`vignette("stablr-python-parity")`.

## Session Info

```{r session-info, eval = has_results}
res$session_info
```
