---
title: "TCGA Breast Cancer: Multi-Omic Biomarker Discovery with stablr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCGA Breast Cancer: Multi-Omic Biomarker Discovery with stablr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 6,
  fig.height = 4,
  cache      = TRUE,
  cache.path = "stablr-tcga-cache/"
)
```



## Overview

This vignette translates the **mixOmics N-Integration Chapter 6** case study
into the `stablr` idiom.  The original mixOmics workflow uses sparse multi-block
Discriminant Analysis (sgccaDA) to classify breast cancer subtypes.  Here we
ask the same scientific question --- which mRNA and miRNA features robustly
distinguish Basal-like tumours from all other subtypes? --- using STABL's
bootstrap stability approach.

**Dataset:** `breast.TCGA` from the `mixOmics` package.

- Training: 150 samples x 520 mRNA features + 150 samples x 184 miRNA features.
- Test: 70 samples x same features.
- Outcome: five PAM50 subtypes, binarised to **Basal vs. non-Basal**.
- Protein data (142 features) is excluded, matching the two-omic scope of
  Chapter 6.

**What you will learn:**

1. How to load and prepare `breast.TCGA` for `stablr`.
2. How to run per-omic STABL as a baseline.
3. How to run the integrated multi-omic pipeline (early + late fusion).
4. How to evaluate validation performance with confusion matrices and BER.
5. How to export results for downstream reporting.

---

## Prerequisites

```{r}
if (!requireNamespace("mixOmics", quietly = TRUE)) {
  stop(
    "This vignette requires mixOmics (Bioconductor). ",
    "Install with: BiocManager::install('mixOmics')"
  )
}
library(stablr)
library(mixOmics)
```



---

## 1. Load and prepare data

`breast.TCGA` is a named list with `$data.train` and `$data.test`, each
containing `$mrna`, `$mirna`, `$protein`, and `$subtype`.

```{r}
data(breast.TCGA)

X_train_mrna  <- breast.TCGA$data.train$mrna
X_train_mirna <- breast.TCGA$data.train$mirna
X_test_mrna   <- breast.TCGA$data.test$mrna
X_test_mirna  <- breast.TCGA$data.test$mirna

cat("Training mRNA:  ", nrow(X_train_mrna),  "x", ncol(X_train_mrna),  "\n")
cat("Training miRNA: ", nrow(X_train_mirna), "x", ncol(X_train_mirna), "\n")
cat("Test mRNA:      ", nrow(X_test_mrna),   "x", ncol(X_test_mrna),   "\n")
cat("Test miRNA:     ", nrow(X_test_mirna),  "x", ncol(X_test_mirna),  "\n")
```



### Binary outcome: Basal vs. non-Basal

```{r}
subtype_train <- breast.TCGA$data.train$subtype
subtype_test  <- breast.TCGA$data.test$subtype

y_train <- as.integer(subtype_train == "Basal")
y_test  <- as.integer(subtype_test  == "Basal")

names(y_train) <- rownames(X_train_mrna)
names(y_test)  <- rownames(X_test_mrna)

cat("Training: ", sum(y_train), "Basal /", sum(y_train == 0), "non-Basal\n")
cat("Test:     ", sum(y_test),  "Basal /", sum(y_test == 0),  "non-Basal\n")
```



---

## 2. Per-omic STABL

We assess each omic independently before running integration.  `n_bootstraps = 50`
is used here for a vignette-friendly runtime; increase to 500-1000 for
publication results.

### 2a. mRNA

```{r}
lambda_mrna <- auto_lambda_grid(
  X_train_mrna, y_train,
  family   = "binomial",
  n_lambda = 20
)
```

```{r}
fit_mrna <- stabl_fit(
  x               = X_train_mrna,
  y               = y_train,
  lambda_grid     = lambda_mrna,
  family          = "binomial",
  n_bootstraps    = 50L,
  artificial_type = "random_permutation",
  random_state    = 42L
)
fit_mrna
```



### 2b. miRNA

```{r}
lambda_mirna <- auto_lambda_grid(
  X_train_mirna, y_train,
  family   = "binomial",
  n_lambda = 20
)
```

```{r}
fit_mirna <- stabl_fit(
  x               = X_train_mirna,
  y               = y_train,
  lambda_grid     = lambda_mirna,
  family          = "binomial",
  n_bootstraps    = 50L,
  artificial_type = "random_permutation",
  random_state    = 42L
)
fit_mirna
```



### Per-omic selected features

```{r}
cat("mRNA selected features:\n")
print(get_support(fit_mrna))

cat("\nmiRNA selected features:\n")
print(get_support(fit_mirna))
```



### Stability paths

```{r}
plot_stabl_path(fit_mrna,  title = "mRNA - stability path")
plot_stabl_path(fit_mirna, title = "miRNA - stability path")
```



---

## 3. Integrated multi-omic pipeline

```{r}
lambda_list <- list(
  mrna  = lambda_mrna,
  mirna = lambda_mirna
)
```

```{r}
multi_fit <- stabl_multiomic_train_validate(
  x_train_list    = list(mrna = X_train_mrna,  mirna = X_train_mirna),
  x_valid_list    = list(mrna = X_test_mrna,   mirna = X_test_mirna),
  y_train         = y_train,
  y_valid         = y_test,
  lambda_grid     = lambda_list,
  family          = "binomial",
  n_bootstraps    = 50L,
  artificial_type = "random_permutation",
  random_state    = 42L,
  early_fusion    = TRUE,
  late_fusion     = TRUE
)
multi_fit
```



---

## 4. Results exploration

### Per-omic biomarkers from the integrated fit

```{r}
cat("mRNA (integrated):\n")
print(get_support(multi_fit$fits$mrna))

cat("\nmiRNA (integrated):\n")
print(get_support(multi_fit$fits$mirna))
```



### Early fusion

```{r}
cat("Early fusion selected features:\n")
print(get_support(multi_fit$early_fusion$fit))
```



### Late fusion weights

```{r}
cat("Late fusion omic weights:\n")
print(multi_fit$late_fusion$weights)
```



### Stability paths from integrated fit

```{r}
plot_stabl_path(multi_fit$fits$mrna,  title = "Integrated - mRNA")
plot_stabl_path(multi_fit$fits$mirna, title = "Integrated - miRNA")
```



---

## 5. Validation performance

```{r}
lf_preds <- multi_fit$late_fusion$valid_predictions

plot_roc(y_true = y_test, y_preds = lf_preds,
         title = "Late fusion - ROC curve (test set)")
plot_prc(y_true = y_test, y_preds = lf_preds,
         title = "Late fusion - PRC curve (test set)")
```



### Confusion matrix and balanced error rate (BER)

```{r}
pred_class <- as.integer(lf_preds >= 0.5)
conf_mat   <- table(Predicted = pred_class, Actual = y_test)
print(conf_mat)

sensitivity <- conf_mat["1", "1"] / sum(conf_mat[, "1"])
specificity <- conf_mat["0", "0"] / sum(conf_mat[, "0"])
ber <- 1 - (sensitivity + specificity) / 2

cat(sprintf("\nSensitivity (Basal):     %.3f\n", sensitivity))
cat(sprintf("Specificity (non-Basal): %.3f\n", specificity))
cat(sprintf("Balanced Error Rate:     %.3f\n", ber))
```



### Feature distribution plots

```{r}
mrna_feats <- names(which(get_support(multi_fit$fits$mrna)))
if (length(mrna_feats) > 0) {
  boxplot_features(
    features = mrna_feats,
    x        = X_train_mrna,
    y        = y_train,
    title    = "Top mRNA features - training set",
    ncol     = min(3L, length(mrna_feats))
  )
}
```



---

## 6. Export results

```{r}
export_stabl_to_csv(multi_fit$fits$mrna,  path = "tcga_results/mrna")
export_stabl_to_csv(multi_fit$fits$mirna, path = "tcga_results/mirna")
```



`export_stabl_to_csv()` writes per-call:

- `<prefix>_stability_scores.csv` - all features with stability scores, sorted.
- `<prefix>_selected_features.csv` - features passing the threshold with FDP.

---

## 7. Comparison with mixOmics

| Aspect | mixOmics (sgccaDA) | stablr |
|---|---|---|
| Method | Sparse CCA + LDA | Bootstrap-stability LASSO |
| Feature selection | L1 penalty tuned by CV | FDP-controlled stability threshold |
| Integration | Joint latent-space optimisation | Early or late fusion |
| Primary metric | BER on test set | FDP+ stability score |
| Output | Components + loadings | Feature names + stability scores |

Both approaches identify a compact, reliable set of mRNA and miRNA biomarkers
for the Basal subtype.  STABL's advantage is explicit FDP control with no
assumption of a shared latent space between omics.

---

## Next steps

- See `vignette("stablr-multiomic")` for a regression multi-omic example.
- See `vignette("stablr-python-parity")` for a Python vs. R result comparison.
- Increase `n_bootstraps` to 500-1000 for publication-quality results.
- Try `artificial_type = "knockoff"` for stronger FDP control.

```{r}
sessionInfo()
```