---
title: "Getting Started with stablr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with stablr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
set.seed(42)
```

## What is STABL?

STABL (Stability-based Thresholding and Aggregation for Biomarker discovery via
LASSO) is a bootstrap-based method for reliable sparse feature selection in
high-dimensional datasets. It runs many bootstrap resamples of a penalised
learner, computes empirical selection probabilities for each feature, and
controls the false-discovery proportion (FDP) by comparing real features against
artificial decoys.

The `stablr` package provides a pure-R implementation with full
`glmnet`-ecosystem compatibility.

**Core idea in one sentence:** a feature is considered a reliable biomarker if
it is selected by the sparse learner *consistently* across many bootstrap
resamples, and its selection frequency is substantially higher than what
randomly permuted noise features achieve.

## Installation

```{r install, eval = FALSE}
# From the repository root after cloning:
devtools::install("r-pkg/stablr")
```

```{r load}
library(stablr)
```

## Learning objectives

By the end of this vignette you will be able to:

1. Build a penalty grid with `auto_lambda_grid()` and understand why it matters.
2. Fit a STABL model with `stabl_fit()` and read the printed summary.
3. Extract selected features (`get_support()`) and ranked stability scores
   (`get_importances()`).
4. Interpret a stability-path plot and an FDR graph.
5. Switch between lasso, adaptive lasso, and elastic-net base learners.

---

## 1. Binary classification

### Simulated data

We generate a 150-sample x 30-feature dataset.  Five features (F1-F5) carry a
true signal; the remaining 25 are pure noise.  Because we know the ground truth,
we can verify that STABL recovers exactly those five features --- making this a
useful sanity check before applying the method to real data.

```{r sim-binary}
n <- 150; p <- 30; n_signal <- 5
X_bin <- matrix(rnorm(n * p), nrow = n,
                dimnames = list(paste0("s", seq_len(n)),
                                paste0("F", seq_len(p))))

# True signal in the first 5 features
beta_bin       <- c(rep(1.5, n_signal), rep(0, p - n_signal))
log_odds       <- X_bin %*% beta_bin
y_bin          <- rbinom(n, size = 1, prob = plogis(log_odds))
names(y_bin)   <- rownames(X_bin)
table(y_bin)
```

The `table()` output shows the class balance of the outcome.  A roughly 50/50
split is ideal; large imbalances can affect how the lasso selects features.

### Fitting STABL

**Step 1 --- build a penalty grid.**  `auto_lambda_grid()` uses `glmnet` to
find the data-dependent range of lambda values and returns a two-column data
frame with columns `lambda` and `alpha`.  Every row becomes one (lambda, alpha)
configuration evaluated during bootstrapping.

**Step 2 --- run the bootstrap loop.**  `stabl_fit()` resamples the data
`n_bootstraps` times, fits the penalised learner at each (lambda, alpha) point,
and records which features are selected.  Key arguments:

| Argument | What it does |
|---|---|
| `family` | Passed to `glmnet`; `"binomial"` for 0/1 outcome, `"gaussian"` for continuous. |
| `n_bootstraps` | Number of bootstrap resamples.  100 is fine for exploration; use 500+ for final results. |
| `artificial_type` | `"random_permutation"` adds column-permuted decoys; `"knockoff"` uses model-X knockoffs for stronger FDP control. |
| `random_state` | Integer seed for full reproducibility. |

We use `n_bootstraps = 100` here for a vignette-friendly runtime.

```{r fit-binary}
lambda_bin <- auto_lambda_grid(X_bin, y_bin, family = "binomial", n_lambda = 20)

fit_bin <- stabl_fit(
  x               = X_bin,
  y               = y_bin,
  lambda_grid     = lambda_bin,
  family          = "binomial",
  n_bootstraps    = 100L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

fit_bin
```

**Reading the printed summary** --- the output reports three key numbers:

- **Stability threshold** --- the minimum selection frequency a feature must
  reach to be called a biomarker (e.g. `0.7` means selected in >= 70% of
  bootstraps).
- **n selected** --- features passing the threshold; for our 5-signal data you
  expect a small number close to 5.
- **FDP estimate** --- estimated upper bound on the false-discovery proportion
  among selected features; values well below 0.05 indicate reliable selection.

### Inspecting results

`get_support()` returns the *names* of features that exceed the stability
threshold --- these are your candidate biomarkers.  On our simulated data you
should see F1-F5 (and ideally nothing else).

`get_importances()` returns a named numeric vector of stability scores
(selection frequencies) for *all* features, sorted from highest to lowest.
A score of 0.8 means the feature was selected in 80% of bootstraps.  The gap
between the top entries and the lower ones shows how cleanly signal features
separate from noise.

```{r results-binary}
# Features selected at the STABL threshold
get_support(fit_bin)

# Stability scores for all features (sorted descending)
head(get_importances(fit_bin), 10)
```

### Visualisation

**Stability path** --- each curve traces one feature's selection frequency as
lambda increases (left = weak penalty, many features selected; right = strong
penalty, few features selected).  Features are coloured by whether they exceed
the threshold.  Look for a visible gap between the top-scoring cluster and the
bulk of near-zero lines: that gap is your signal-to-noise contrast.  On the
simulated data F1-F5 should form a distinct upper band.

```{r plot-binary, fig.alt = "Stability path for binary classification"}
plot_stabl_path(fit_bin, title = "Binary classification - stability path")
```

**FDR graph** --- the x-axis sweeps candidate thresholds from 0 to 1; the
y-axis shows the estimated FDP at each threshold.  The dashed vertical line
marks the chosen threshold (minimises FDP while keeping at least one feature);
the horizontal dashed line is the FDP target (default 0.05).  A good result
shows the chosen threshold sitting below the FDP target line.

```{r fdr-binary, fig.alt = "FDR estimate graph for binary classification"}
plot_fdr_graph(fit_bin, title = "Binary classification - FDR estimate")
```

---

## 2. Regression

### Simulated data

We reuse the same 150 x 30 dimensions but draw an independent feature matrix
and generate a *continuous* outcome (linear combination of F1-F5 plus Gaussian
noise).  The only change to the fitting workflow is `family = "gaussian"`.

```{r sim-reg}
X_reg <- matrix(rnorm(n * p), nrow = n,
                dimnames = list(paste0("s", seq_len(n)),
                                paste0("F", seq_len(p))))
beta_reg     <- c(rep(2, n_signal), rep(0, p - n_signal))
y_reg        <- X_reg %*% beta_reg + rnorm(n, sd = 1)
names(y_reg) <- rownames(X_reg)
```

### Fitting STABL

```{r fit-reg}
lambda_reg <- auto_lambda_grid(X_reg, y_reg, family = "gaussian", n_lambda = 20)

fit_reg <- stabl_fit(
  x               = X_reg,
  y               = y_reg,
  lambda_grid     = lambda_reg,
  family          = "gaussian",
  n_bootstraps    = 100L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

fit_reg
```

### Results

`get_support()` should again return F1-F5.  You may notice slightly different
stability scores compared to the binary fit --- this is expected because the
Gaussian loss function changes the effective sparsity of each bootstrap model.

```{r results-reg}
get_support(fit_reg)
head(get_importances(fit_reg), 10)
```

### Visualisation

The stability path should look qualitatively similar to the binary case: F1-F5
curve upward and separate from the noise features.  Small differences in curve
heights between the two tasks reflect how each family amplifies feature signal.

```{r plot-reg, fig.alt = "Stability path for regression"}
plot_stabl_path(fit_reg, title = "Regression - stability path")
```

---

## 3. Learner variants

### Adaptive lasso

The adaptive lasso re-weights the L1 penalty so that features with small
coefficients in a preliminary ridge fit are penalised more heavily.  This
sharpens the stability contrast: true signal features accumulate higher
selection frequencies while noise features are pushed further toward zero.

**When to use it:** prefer adaptive lasso when you expect a sparse true signal
and want a tighter biomarker list with fewer borderline features.

**What to expect:** `get_support(fit_ada)` should return F1-F5, and
`get_importances()` will typically show higher scores for true features
compared to plain lasso.

```{r adaptive-lasso}
lambda_ada <- auto_lambda_grid(X_bin, y_bin, family = "binomial", n_lambda = 20)

fit_ada <- stabl_fit(
  x               = X_bin,
  y               = y_bin,
  lambda_grid     = lambda_ada,
  base_learner    = "adaptive_lasso",
  family          = "binomial",
  n_bootstraps    = 100L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

get_support(fit_ada)
```

### Elastic net

Elastic net interpolates between lasso (alpha = 1) and ridge (alpha = 0).
The L2 component encourages correlated features to be selected together, which
is useful when biomarkers form functional modules (e.g. co-expressed genes).

Passing multiple `l1_ratio` values to `auto_lambda_grid()` makes STABL sweep
the (lambda, alpha) grid jointly, exploring both the sparsity level and the
L1/L2 mix in every bootstrap.

**When to use it:** prefer elastic net over plain lasso when features are highly
correlated and you want to avoid arbitrarily picking one representative from a
correlated group.

```{r elastic-net}
lambda_en <- auto_lambda_grid(
  X_bin, y_bin,
  family = "binomial", n_lambda = 10,
  l1_ratio = c(0.5, 0.9)
)

fit_en <- stabl_fit(
  x               = X_bin,
  y               = y_bin,
  lambda_grid     = lambda_en,
  base_learner    = "elastic_net",
  family          = "binomial",
  n_bootstraps    = 100L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

get_support(fit_en)
```

---

## Next steps

- See `vignette("stablr-multiomic")` for multi-omic integration with early and
  late fusion on a real Onset of Labor dataset.
- See `?stabl_multiomic_cv` for grouped cross-validation workflows.
- See `?save_stabl_results` for exporting scores, plots, and selected-feature
  tables to disk.
